---
title: "MASDS Thesis"
author: "Lauren Huang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(utils)
library(dplyr)
library(purrr)
library(fs)
library(stringr)
library(ggplot2)
library(sf)
library(tmap)
library(cartogram)
library(stelfi)
library(lubridate)
library(hawkes)
library(MASS)
library(fitdistrplus)
library(PtProcess)
library(stats)
library(foreach)
library(sn)
library(geosphere)
library(spData)
library(sf)
library(tidyr)
library(geodist)
library(vegan)
library(ade4)
library(scales)
library(patchwork)
library(parallel)
library(reshape2)
```

# Read in the Data

```{r}
# parent directory containing relevant folders
parent_dir <- "C:/Users/cupca/Desktop/UCLA/Thesis"

# all folders within the parent directory
folder_list <- dir_ls(path = parent_dir, type = "dir")

# folders within the parent directory
folders <- c("2020", "2021", "2022", "2023")

# construct full paths to the specified directories
full_paths <- file.path(parent_dir, folders)

# a function to read and standardize columns of each csv file
read_and_standardize <- function(file) {
  df <- read.csv(file)
  # convert certain columns to the same format
  columns_to_convert <- c("Cases", "Case.Rate")
  # replace occurrences of -- with 0
  value_to_replace <- "--"
  # convert occurrences of the specific character value in the specified columns to integer 0
  df <- df %>%
  mutate(across(all_of(columns_to_convert), 
                 ~ ifelse(. == value_to_replace, 0, .))) %>%
  mutate(across(all_of(columns_to_convert), 
                 ~ as.numeric(as.character(.))))
  
  # standardize columns: add NA/missing columns if they do not exist
  if (!("Deaths" %in% names(df))) {
    df$Deaths <- NA
  }
  if (!("Death.Rate" %in% names(df))) {
    df$Death.Rate <- NA
  }
  return(df)
}

# read, combine all csv files from all folders
all_covid_data <- full_paths %>%
  # list all csv files in the folder
  map(~ dir_ls(path = .x, regexp = "\\.csv$", type = "file")) %>%
  # flatten list of file paths
  unlist() %>%
  # read, standardize each csv file
  map_df(read_and_standardize, .id = "source_id")

# save combined files to a new csv file
write.csv(all_covid_data, "all_covid_data.csv", row.names = FALSE)
```

## Data Cleaning

```{r}
# define a function that relabels source_id as Year
replace_if_contains <- function(df, substring_to_find, new_value) {
  # replace values containing the substring in the specified column (source_id)
  df <- df %>%
    mutate(source_id = ifelse(str_detect(source_id, substring_to_find), new_value, source_id))
  return(df)
}

all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2020", new_value="2020")
all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2021", new_value="2021")
all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2022", new_value="2022")
all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2023", new_value="2023")

# rename column as Year
names(all_covid_data)[names(all_covid_data) == 'source_id'] <- 'year'

# Date Preprocessing

# extract month
all_covid_data$month <- substr(all_covid_data$date, nchar(all_covid_data$date) - 3, nchar(all_covid_data$date) - 2)

# extract day
all_covid_data$day <- substr(all_covid_data$date, nchar(all_covid_data$date) - 1, nchar(all_covid_data$date))

# create a column containing the full date (month, day, year)
all_covid_data$date <- as.Date(paste(all_covid_data$month, all_covid_data$day, all_covid_data$year), "%m%d%Y")

# drop individual month, day, year columns we no longer need
all_covid_data <- subset(all_covid_data, select = -c(year, month, day))

# create new column called incidence, counting new deaths per day in each community
all_covid_data <- all_covid_data %>%
  arrange(name, date) %>%
  group_by(name) %>%
  mutate(incidence = Deaths - lag(Deaths, default = 0)) %>%
  ungroup()

# replace negative or NA values in the incidence column with 0
all_covid_data <- all_covid_data %>%
  mutate(incidence = ifelse(is.na(incidence) | incidence < 0, 0, incidence))

# replace NA values with 0 for incidence
all_covid_data <- all_covid_data %>%
  mutate(Deaths = ifelse(is.na(Deaths), 0, Deaths))

# convert the date column to Date type
community_data <- all_covid_data %>%
  mutate(date = as.Date(date)) %>%
  filter(date > 0) %>%
  filter(('2020-05-16' < date) & (date < '2023-01-02'))

# replace NA, 0 cases with 0.5 or 1
community_data <- community_data %>%
  mutate(Deaths = ifelse(is.na(Deaths), 0.5, Deaths),
         Deaths = ifelse(Deaths == 0, 0.5, Deaths))

# convert dates to numeric time (days since the start of the data)
start_date <- min(community_data$date, na.rm = TRUE)
community_data <- community_data %>%
  mutate(time = as.numeric(difftime(date, start_date, units = "days")))

# remove rows with invalid times
community_data <- community_data %>%
  filter(!is.na(time))
```

## Daily death incidence

prepare daily incidence data

```{r}
# Santa Monica
santa_monica <- community_data %>%
  filter(name == "City of Santa Monica")

# Silverlake
silverlake <- community_data %>%
  filter(name == "Los Angeles - Silverlake")

# Pomona
pomona <- community_data %>%
  filter(name == "City of Pomona")

# Glendale
glendale <- community_data %>%
  filter(name == "City of Glendale")

# El Monte
el_monte <- community_data %>%
  filter(name == "City of El Monte")

# Little Armenia
little_armenia <- community_data %>%
  filter(name == "Los Angeles - Little Armenia")

# Signal Hill
signal_hill <- community_data %>%
  filter(name == "City of Signal Hill")
```

prepare transformed event times for each community

```{r}
# cannot be simultaneous events, need to transform event times
generate_event_times <- function(time, deaths) {
  if (deaths > 0) {
    # distribute event times uniformly within the same day
    return(time + runif(deaths, min = 0, max = 1))  # random times
  } else {
    return(NULL)  # no deaths on this day
  }
}

# generate event times for each community
event_times <- community_data %>%
  group_by(name) %>%
  summarise(event_times = list(unlist(mapply(generate_event_times, time, incidence))))
# using death incidence instead of cumulative deaths

# Santa Monica
event_times_santa_monica <- event_times %>%
  filter(name == "City of Santa Monica") %>%
  pull(event_times)
# sort and get unique times
event_times_santa_monica <- sort(unlist(event_times_santa_monica))
event_times_santa_monica_unique <- unique(event_times_santa_monica)
adj_event_times_santa_monica <- event_times_santa_monica_unique + 
  seq(0, length(event_times_santa_monica_unique) - 1) * 1e-5

# Silverlake
event_times_silverlake <- event_times %>%
  filter(name == "Los Angeles - Silverlake") %>%
  pull(event_times)
# sort and get unique times
event_times_silverlake <- sort(unlist(event_times_silverlake))
event_times_silverlake_unique <- unique(event_times_silverlake)
adj_event_times_silverlake <- event_times_silverlake_unique + 
  seq(0, length(event_times_silverlake_unique) - 1) * 1e-5


# Pomona
event_times_pomona <- event_times %>%
  filter(name == "City of Pomona") %>%
  pull(event_times)
# sort and get unique times
event_times_pomona <- sort(unlist(event_times_pomona))
event_times_pomona_unique <- unique(event_times_pomona)
adj_event_times_pomona <- event_times_pomona_unique + 
  seq(0, length(event_times_pomona_unique) - 1) * 1e-5


# Glendale
event_times_glendale <- event_times %>%
  filter(name == "City of Glendale") %>%
  pull(event_times)
# sort and get unique times
event_times_glendale <- sort(unlist(event_times_glendale))
event_times_glendale_unique <- unique(event_times_glendale)
adj_event_times_glendale <- event_times_glendale_unique + 
  seq(0, length(event_times_glendale_unique) - 1) * 1e-5


# El Monte
event_times_el_monte <- event_times %>%
  filter(name == "City of El Monte") %>%
  pull(event_times)
# sort and get unique times
event_times_el_monte <- sort(unlist(event_times_el_monte))
event_times_el_monte_unique <- unique(event_times_el_monte)
adj_event_times_el_monte <- event_times_el_monte_unique + 
  seq(0, length(event_times_el_monte_unique) - 1) * 1e-5

# Little Armenia
event_times_little_armenia <- event_times %>%
  filter(name == "Los Angeles - Little Armenia") %>%
  pull(event_times)
# sort and get unique times
event_times_little_armenia <- sort(unlist(event_times_little_armenia))
event_times_little_armenia_unique <- unique(event_times_little_armenia)
adj_event_times_little_armenia <- event_times_little_armenia_unique + 
  seq(0, length(event_times_little_armenia_unique) - 1) * 1e-5

# Signal Hill
event_times_signal_hill <- event_times %>%
  filter(name == "City of Signal Hill") %>%
  pull(event_times)
# sort and get unique times
event_times_signal_hill <- sort(unlist(event_times_signal_hill))
event_times_signal_hill_unique <- unique(event_times_signal_hill)
adj_event_times_signal_hill <- event_times_signal_hill_unique + 
  seq(0, length(event_times_signal_hill_unique) - 1) * 1e-5
```

## Weekly death incidence

prepare weekly incidence data

```{r}
# add a 'week' column representing the start of the week
all_covid_weekly <- all_covid_data %>%
  mutate(week = floor_date(date, unit = "week"))

# aggregate the data by community and week
all_covid_weekly <- all_covid_weekly %>%
  group_by(name, week) %>%
  summarise(
    weekly_cases = sum(Cases, na.rm = TRUE), # sum cases each week
    weekly_deaths = sum(Deaths, na.rm = TRUE), # sum deaths each week
    weekly_incidence = sum(incidence, na.rm = TRUE), # sum death incidence each week
    population = first(population) # use the first pop value
  ) %>%
  ungroup()

# calculate derived metrics (ex. weekly death rate)
all_covid_weekly <- all_covid_weekly %>%
  mutate(weekly_death_rate = (weekly_deaths / population) * 100000, # death rate per 100k
  weekly_incidence = ifelse(is.na(weekly_incidence), 0, weekly_incidence)) # replace NA values with 0

# convert the date column to Date type
community_data_weekly <- all_covid_weekly %>%
  mutate(week = as.Date(week)) %>%
  filter(week > 0) %>%
  filter(('2020-05-16' < week) & (week < '2023-01-02')) %>% 
  # exclude date gap 11/21/22-1/2/23 and 1/3/23-11/21/23
  mutate(weekly_deaths = ifelse(is.na(weekly_deaths), 0.5, weekly_deaths), # replace NA, 0 with 0.5
         weekly_deaths = ifelse(weekly_deaths == 0, 0.5, weekly_deaths),
         weekly_incidence = ifelse(is.na(weekly_incidence), 0.5, weekly_incidence),
         weekly_incidence = ifelse(weekly_incidence == 0, 0.5, weekly_incidence))

# convert dates to numeric time (weeks since the start)
start_week <- min(community_data_weekly$week, na.rm = TRUE)
community_data_weekly <- community_data_weekly %>%
  mutate(time = as.numeric(difftime(week, start_week, units = "weeks")))

# look for NA values, remove rows with invalid times
community_data_weekly <- community_data_weekly %>%
  filter(!is.na(time))
```

```{r}
# convert dates to weekly scale
santa_monica_weekly <- santa_monica %>%
  mutate(week = floor_date(date, unit = "week")) %>%
  group_by(week) %>%
  summarize(
    weekly_incidence = sum(incidence, na.rm = TRUE),   # sum deaths per week
    total_population = first(population),              # assume population is constant
    total_deaths = sum(Deaths, na.rm = TRUE),          # sum deaths per week
    avg_death_rate = mean(Death.Rate, na.rm = TRUE),   # avg death rate per week
    avg_cases = mean(Cases, na.rm = TRUE),             # avg cases per week
    avg_case_rate = mean(Case.Rate, na.rm = TRUE),     # avg case rate per week
    .groups = "drop"
  ) %>%
  filter(week > as.Date("2020-05-16") & week < as.Date("2023-01-02"))

silverlake_weekly <- silverlake %>%
  mutate(week = floor_date(date, unit = "week")) %>%
  group_by(week) %>%
  summarize(
    weekly_incidence = sum(incidence, na.rm = TRUE),
    total_population = first(population),
    total_deaths = sum(Deaths, na.rm = TRUE),
    avg_death_rate = mean(Death.Rate, na.rm = TRUE),
    avg_cases = mean(Cases, na.rm = TRUE),
    avg_case_rate = mean(Case.Rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(week > as.Date("2020-05-16") & week < as.Date("2023-01-02"))

pomona_weekly <- pomona %>%
  mutate(week = floor_date(date, unit = "week")) %>%
  group_by(week) %>%
  summarize(
    weekly_incidence = sum(incidence, na.rm = TRUE),
    total_population = first(population),
    total_deaths = sum(Deaths, na.rm = TRUE),
    avg_death_rate = mean(Death.Rate, na.rm = TRUE),
    avg_cases = mean(Cases, na.rm = TRUE),
    avg_case_rate = mean(Case.Rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(week > as.Date("2020-05-16") & week < as.Date("2023-01-02"))

glendale_weekly <- glendale %>%
  mutate(week = floor_date(date, unit = "week")) %>%
  group_by(week) %>%
  summarize(
    weekly_incidence = sum(incidence, na.rm = TRUE),
    total_population = first(population),
    total_deaths = sum(Deaths, na.rm = TRUE),
    avg_death_rate = mean(Death.Rate, na.rm = TRUE),
    avg_cases = mean(Cases, na.rm = TRUE),
    avg_case_rate = mean(Case.Rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(week > as.Date("2020-05-16") & week < as.Date("2023-01-02"))

el_monte_weekly <- el_monte %>%
  mutate(week = floor_date(date, unit = "week")) %>%
  group_by(week) %>%
  summarize(
    weekly_incidence = sum(incidence, na.rm = TRUE),
    total_population = first(population),
    total_deaths = sum(Deaths, na.rm = TRUE),
    avg_death_rate = mean(Death.Rate, na.rm = TRUE),
    avg_cases = mean(Cases, na.rm = TRUE),
    avg_case_rate = mean(Case.Rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(week > as.Date("2020-05-16") & week < as.Date("2023-01-02"))

little_armenia_weekly <- little_armenia %>%
  mutate(week = floor_date(date, unit = "week")) %>%
  group_by(week) %>%
  summarize(
    weekly_incidence = sum(incidence, na.rm = TRUE),
    total_population = first(population),
    total_deaths = sum(Deaths, na.rm = TRUE),
    avg_death_rate = mean(Death.Rate, na.rm = TRUE),
    avg_cases = mean(Cases, na.rm = TRUE),
    avg_case_rate = mean(Case.Rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(week > as.Date("2020-05-16") & week < as.Date("2023-01-02"))

signal_hill_weekly <- signal_hill %>%
  mutate(week = floor_date(date, unit = "week")) %>%
  group_by(week) %>%
  summarize(
    weekly_incidence = sum(incidence, na.rm = TRUE),
    total_population = first(population), 
    total_deaths = sum(Deaths, na.rm = TRUE),
    avg_death_rate = mean(Death.Rate, na.rm = TRUE),
    avg_cases = mean(Cases, na.rm = TRUE),
    avg_case_rate = mean(Case.Rate, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  filter(week > as.Date("2020-05-16") & week < as.Date("2023-01-02"))
```

prepare transformed event times for each community

```{r}
# cannot be "simultaneous events" (ex. multiple deaths on a day), need to transform event times
generate_event_times <- function(time, deaths) {
  if (deaths > 0) {
    # distribute event times uniformly within the same day
    return(time + runif(deaths, min = 0, max = 1))  # random times within the day
  } else {
    return(NULL)  # no deaths on this day
  }
}

# generate event times for each community
event_times_weekly <- community_data_weekly %>%
  group_by(name) %>%
  summarise(event_times = list(unlist(mapply(generate_event_times, time, weekly_incidence)))) 

# Santa Monica
event_times_santa_monica_weekly <- event_times_weekly %>%
  filter(name == "City of Santa Monica") %>%
  pull(event_times)
# sort and get unique times
event_times_santa_monica_weekly <- sort(unlist(event_times_santa_monica_weekly))
event_times_santa_monica_weekly_unique <- unique(event_times_santa_monica_weekly)
adj_event_times_santa_monica_weekly <- event_times_santa_monica_weekly_unique + 
  seq(0, length(event_times_santa_monica_weekly_unique) - 1) * 1e-5

# Silverlake
event_times_silverlake_weekly <- event_times_weekly %>%
  filter(name == "Los Angeles - Silverlake") %>%
  pull(event_times)
# sort and get unique times
event_times_silverlake_weekly <- sort(unlist(event_times_silverlake_weekly))
event_times_silverlake_weekly_unique <- unique(event_times_silverlake_weekly)
adj_event_times_silverlake_weekly <- event_times_silverlake_weekly_unique + 
  seq(0, length(event_times_silverlake_weekly_unique) - 1) * 1e-5

# Pomona
event_times_pomona_weekly <- event_times_weekly %>%
  filter(name == "City of Pomona") %>%
  pull(event_times)
# sort and get unique times
event_times_pomona_weekly <- sort(unlist(event_times_pomona_weekly))
event_times_pomona_weekly_unique <- unique(event_times_pomona_weekly)
adj_event_times_pomona_weekly <- event_times_pomona_weekly_unique + 
  seq(0, length(event_times_pomona_weekly_unique) - 1) * 1e-5

# Glendale
event_times_glendale_weekly <- event_times_weekly %>%
  filter(name == "City of Glendale") %>%
  pull(event_times)
# sort and get unique times
event_times_glendale_weekly <- sort(unlist(event_times_glendale_weekly))
event_times_glendale_weekly_unique <- unique(event_times_glendale_weekly)
adj_event_times_glendale_weekly <- event_times_glendale_weekly_unique + 
  seq(0, length(event_times_glendale_weekly_unique) - 1) * 1e-5

# El Monte
event_times_el_monte_weekly <- event_times_weekly %>%
  filter(name == "City of El Monte") %>%
  pull(event_times)
# sort and get unique times
event_times_el_monte_weekly <- sort(unlist(event_times_el_monte_weekly))
event_times_el_monte_weekly_unique <- unique(event_times_el_monte_weekly)
adj_event_times_el_monte_weekly <- event_times_el_monte_weekly_unique + 
  seq(0, length(event_times_el_monte_weekly_unique) - 1) * 1e-5

# Little Armenia
event_times_little_armenia_weekly <- event_times_weekly %>%
  filter(name == "Los Angeles - Little Armenia") %>%
  pull(event_times)
# sort and get unique times
event_times_little_armenia_weekly <- sort(unlist(event_times_little_armenia_weekly))
event_times_little_armenia_weekly_unique <- unique(event_times_little_armenia_weekly)
adj_event_times_little_armenia_weekly <- event_times_little_armenia_weekly_unique + 
  seq(0, length(event_times_little_armenia_weekly_unique) - 1) * 1e-5

# Signal Hill
event_times_signal_hill_weekly <- event_times_weekly %>%
  filter(name == "City of Signal Hill") %>%
  pull(event_times)
# sort and get unique times
event_times_signal_hill_weekly <- sort(unlist(event_times_signal_hill_weekly))
event_times_signal_hill_weekly_unique <- unique(event_times_signal_hill_weekly)
adj_event_times_signal_hill_weekly <- event_times_signal_hill_weekly_unique + 
  seq(0, length(event_times_signal_hill_weekly_unique) - 1) * 1e-5
```

# Hawkes Modeling and Analysis

## Hawkes Model Process using stelfi package

Note: Code below adapted from tutorial in the R package stelfi documentation: https://cmjt.github.io/stelfi/univariate-hawkes.html

How to interpret show_hawkes() output
1. top: the estimated conditional intensity
2. bottom: the data

How to interpret GOF output plots
1. top left, plots the compensator values against the observed times, which under a well fitting model should align;
2. top right, a transformed QQ plot, the observed quantities should align with the theoretical quantiles under a well fitting model;
3. bottom left, the compensator differences, which under the model are assumed to be distributed;
4. bottom right, consecutive compensator differences, which should show no obvious pattern (no autocorrelation evident) under a well fitting model.

### stelfi daily

```{r}
# function to process and plot
stelfi_daily <- function(community_name, community_data, event_times, start_values) {
  
  # adjust event times
  event_times <- sort(unlist(event_times))
  event_times_unique <- unique(event_times)
  adj_event_times <- event_times_unique + seq(0, length(event_times_unique) - 1) * 1e-5
  
  # fit model
  fit <- fit_hawkes(times = adj_event_times, parameters = start_values)
  
  # model summary and gof plots
  print(show_hawkes(fit))
  print(show_hawkes_GOF(fit))
  
  # get parameters
  params <- get_coefs(fit)
  mu <- params[1]
  alpha <- params[2]
  beta <- params[3]
  
  # calculate expected deaths
  compute_expected_deaths <- function(event_times, mu, alpha, beta) {
    days <- max(event_times)
    lambda_t <- numeric(days)
    for (t in 1:days) {
      past_events <- event_times[event_times < t]
      lambda_t[t] <- mu + sum(alpha * exp(-beta * (t - past_events)))
    }
    return(lambda_t)
  }
  
  expected_deaths <- compute_expected_deaths(adj_event_times, mu, alpha, beta)
  expected_deaths <- expected_deaths[1:794]  # daily event times range
  
  # create date sequence
  start_date <- min(community_data$date)
  date_sequence <- seq(start_date, by = "day", length.out = 794)
  
  df <- data.frame(
    date = community_data$date,
    observed = community_data$incidence,
    expected = expected_deaths
  )
  
  # plot
  plot <- ggplot(df, aes(x = date)) +
    geom_line(aes(y = observed, color = "Observed"), linewidth = 1) +
    geom_line(aes(y = expected, color = "Expected"), linewidth = 1) +
    labs(title = paste("Expected vs. Observed Deaths for", community_name),
         x = "Date", y = "Daily Deaths") +
    scale_color_manual(values = c("red", "black")) +
    theme_minimal()
  
  return(list(
    expected_deaths = expected_deaths,
    adj_event_times = adj_event_times,
    plot = plot
  ))
}

results_daily_sm <- stelfi_daily("Santa Monica", santa_monica, event_times_santa_monica, 
                                c(mu = 9, alpha = 3, beta = 10))
results_daily_s  <- stelfi_daily("Silverlake", silverlake, event_times_silverlake, 
                                c(mu = 9, alpha = 3, beta = 10))
results_daily_p  <- stelfi_daily("Pomona", pomona, event_times_pomona, 
                                c(mu = 9, alpha = 3, beta = 10))
results_daily_g  <- stelfi_daily("Glendale", glendale, event_times_glendale, 
                                c(mu = 9, alpha = 3, beta = 10))
results_daily_em <- stelfi_daily("El Monte", el_monte, event_times_el_monte, 
                                c(mu = 9, alpha = 3, beta = 10))
results_daily_la <- stelfi_daily("Little Armenia", little_armenia, event_times_little_armenia, 
                                c(mu = 9, alpha = 3, beta = 10))
results_daily_sh <- stelfi_daily("Signal Hill", signal_hill, event_times_signal_hill, 
                                c(mu = 9, alpha = 3, beta = 10))
```

```{r}
# combine plots
(results_daily_sm$plot | results_daily_s$plot | results_daily_p$plot) / 
  (results_daily_g$plot | results_daily_em$plot | results_daily_la$plot) / 
  (results_daily_sh$plot)

results_daily_sm$plot

(results_daily_s$plot | results_daily_p$plot | results_daily_g$plot) /
  (results_daily_em$plot | results_daily_la$plot | results_daily_sh$plot)
```

RMSE & MAE

```{r}
# observed deaths
observed_deaths_santa_monica <- santa_monica$incidence
observed_deaths_silverlake <- silverlake$incidence
observed_deaths_pomona <- pomona$incidence
observed_deaths_glendale <- glendale$incidence
observed_deaths_el_monte <- el_monte$incidence
observed_deaths_little_armenia <- little_armenia$incidence
observed_deaths_signal_hill <- signal_hill$incidence

# expected/predicted deaths
expected_deaths_santa_monica <- results_daily_sm$expected_deaths
expected_deaths_silverlake <- results_daily_s$expected_deaths
expected_deaths_pomona <- results_daily_p$expected_deaths
expected_deaths_glendale <- results_daily_g$expected_deaths
expected_deaths_el_monte <- results_daily_em$expected_deaths
expected_deaths_little_armenia <- results_daily_la$expected_deaths
expected_deaths_signal_hill <- results_daily_sh$expected_deaths


observed_deaths <- list(
  santa_monica = observed_deaths_santa_monica,
  silverlake = observed_deaths_silverlake,
  pomona = observed_deaths_pomona,
  glendale = observed_deaths_glendale,
  el_monte = observed_deaths_el_monte,
  little_armenia = observed_deaths_little_armenia,
  signal_hill = observed_deaths_signal_hill
)

expected_deaths <- list(
  santa_monica = expected_deaths_santa_monica,
  silverlake = expected_deaths_silverlake,
  pomona = expected_deaths_pomona,
  glendale = expected_deaths_glendale,
  el_monte = expected_deaths_el_monte,
  little_armenia = expected_deaths_little_armenia,
  signal_hill = expected_deaths_signal_hill
)

# calculate RMSE
calculate_rmse <- function(observed, expected) {
  sqrt(mean((observed - expected)^2, na.rm = TRUE))
}
rmse_values <- sapply(names(observed_deaths), function(community) {
  calculate_rmse(observed_deaths[[community]], expected_deaths[[community]])
})
rmse_daily_stelfi <- data.frame(
  community = names(observed_deaths),
  rmse = rmse_values
)
print(rmse_daily_stelfi)

# calculate MAE
calculate_mae <- function(observed, expected) {
  mean(abs(observed - expected), na.rm = TRUE)
}
mae_values <- sapply(names(observed_deaths), function(community) {
  calculate_mae(observed_deaths[[community]], expected_deaths[[community]])
})
mae_daily_fit <- data.frame(
  community = names(observed_deaths),
  mae = mae_values
)
print(mae_daily_fit)
```

### Likelihood ratio test: Gamma vs Exponential distribution

- $H_0$: Reject the null hypothesis that the data follows an exponential distribution. The gamma model (with $alpha \neq 1$ fits significantly better.
- $H_a$: Fail to reject the null hypothesis. The simpler exponential model is enough.

1. If LRT rejects null hypothesis, run KS Gamma test.
2. If LRT does not reject null hypothesis, run KS Exponential test.

```{r}
# list of communities
communities <- list(
  santa_monica = adj_event_times_santa_monica,
  silverlake = adj_event_times_silverlake,
  pomona = adj_event_times_pomona,
  glendale = adj_event_times_glendale,
  el_monte = adj_event_times_el_monte,
  little_armenia = adj_event_times_little_armenia,
  signal_hill = adj_event_times_signal_hill
)

# Likelihood Ratio Test (Gamma vs. Exponential)
likelihood_ratio_test <- function(event_times, community_name) {
  interarrival_times <- diff(event_times)
  
  gamma_fit <- fitdist(interarrival_times, "gamma")
  exp_fit <- fitdist(interarrival_times, "exp")
  
  loglik_gamma <- sum(dgamma(interarrival_times, shape = gamma_fit$estimate["shape"],
                             rate = gamma_fit$estimate["rate"], log = TRUE))
  
  loglik_exp <- sum(dexp(interarrival_times, rate = exp_fit$estimate["rate"], log = TRUE))
  
  LR_statistic <- -2 * (loglik_exp - loglik_gamma)
  p_value <- pchisq(LR_statistic, df = 1, lower.tail = FALSE)
  
  cat("\n=== Likelihood Ratio Test for Community:", community_name, "===\n")
  cat("Log-likelihood (Gamma):", loglik_gamma, "\n")
  cat("Log-likelihood (Exponential):", loglik_exp, "\n")
  cat("Likelihood Ratio Statistic:", LR_statistic, "\n")
  cat("P-value:", p_value, "\n")
}

# KS Test for Gamma fit
ks_gamma_test <- function(event_times, community_name) {
  inter_event_times <- diff(event_times)
  
  gamma_fit <- fitdist(inter_event_times, "gamma", method = "mle")
  
  shape <- gamma_fit$estimate["shape"]
  rate <- gamma_fit$estimate["rate"]
  
  ks_test <- ks.test(inter_event_times, "pgamma", shape = shape, rate = rate)
  
  cat("\n=== KS Gamma Test for Community:", community_name, "===\n")
  print(ks_test)
}

# KS Test for Exponential fit
ks_exp_test <- function(event_times, community_name) {
  inter_event_times <- diff(event_times)
  
  exp_fit <- fitdist(inter_event_times, "exp", method = "mle")
  
  rate <- exp_fit$estimate["rate"]
  
  ks_test <- ks.test(inter_event_times, "pexp", rate = rate)
  
  cat("\n=== KS Exponential Test for Community:", community_name, "===\n")
  print(ks_test)
}

# loop through communities
for (community in names(communities)) {
  event_times <- communities[[community]]
  
  likelihood_ratio_test(event_times, community)
  ks_gamma_test(event_times, community)
  ks_exp_test(event_times, community)
}
```

### stelfi weekly

Now, run model on the weekly death scale.

```{r}
# function for comparing weekly expected deaths vs observed deaths
stelfi_weekly <- function(community_label, community_data, event_times, start_params) {
  
  # adjust event times
  event_times_sorted <- sort(unlist(event_times))
  event_times_unique <- unique(event_times_sorted)
  adj_event_times <- event_times_unique + seq(0, length(event_times_unique) - 1) * 1e-5
  
  # fit model
  fit <- fit_hawkes(times = adj_event_times, parameters = start_params)
  
  # model summary and gof plots
  print(show_hawkes(fit))
  print(show_hawkes_GOF(fit))
  
  # get parameters
  params <- get_coefs(fit)
  mu <- params[1]
  alpha <- params[2]
  beta <- params[3]
  
  # calculate expected deaths
  compute_expected_deaths <- function(event_times, mu, alpha, beta) {
    days <- max(event_times)
    lambda_t <- numeric(days)
    for (t in 1:days) {
      past_events <- event_times[event_times < t]
      lambda_t[t] <- mu + sum(alpha * exp(-beta * (t - past_events)))
    }
    return(lambda_t)
  }
  
  # calculate expected deaths
  expected_deaths <- compute_expected_deaths(adj_event_times, mu, alpha, beta)
  expected_deaths <- expected_deaths[1:132] # weekly event times range
  
  # prepare observed weekly data
  community_weekly <- community_data %>%
    mutate(week = floor_date(date, unit = "week")) %>%
    group_by(week) %>%
    summarize(weekly_incidence = sum(incidence, na.rm = TRUE)) %>%
    ungroup() %>%
    filter(week > as.Date("2020-05-16") & week < as.Date("2023-01-02"))
  
  df_weekly <- data.frame(
    date = community_weekly$week,
    observed = community_weekly$weekly_incidence,
    expected = expected_deaths
  )
  
  # plot
  plot <- ggplot(df_weekly, aes(x = date)) +
    geom_line(aes(y = observed, color = "Observed"), linewidth = 1) +
    geom_line(aes(y = expected, color = "Expected"), linewidth = 1) +
    labs(title = paste("Expected vs. Observed Deaths for", community_label),
         x = "Date", y = "Weekly Deaths") +
    scale_color_manual(values = c("red", "black")) +
    theme_minimal()
  
  return(list(
    expected_deaths = expected_deaths,
    adj_event_times = adj_event_times,
    plot = plot
  ))
}

results_weekly_sm <- stelfi_weekly("Santa Monica", santa_monica, event_times_santa_monica_weekly, 
                                c(mu = 9, alpha = 3, beta = 10))
results_weekly_s  <- stelfi_weekly("Silverlake", silverlake, event_times_silverlake_weekly, 
                                c(mu = 9, alpha = 3, beta = 10))
results_weekly_p  <- stelfi_weekly("Pomona", pomona, event_times_pomona_weekly, 
                                c(mu = 9, alpha = 3, beta = 10))
results_weekly_g  <- stelfi_weekly("Glendale", glendale, event_times_glendale_weekly, 
                                c(mu = 9, alpha = 3, beta = 10))
results_weekly_em <- stelfi_weekly("El Monte", el_monte, event_times_el_monte_weekly, 
                                c(mu = 9, alpha = 3, beta = 10))
results_weekly_la <- stelfi_weekly("Little Armenia", little_armenia, event_times_little_armenia_weekly, 
                                c(mu = 9, alpha = 3, beta = 10))
results_weekly_sh <- stelfi_weekly("Signal Hill", signal_hill, event_times_signal_hill_weekly, 
                                c(mu = 9, alpha = 3, beta = 10))
```

```{r}
# combine plots
(results_weekly_sm$plot | results_weekly_s$plot | results_weekly_p$plot) / 
  (results_weekly_g$plot | results_weekly_em$plot | results_weekly_la$plot) / 
  (results_weekly_sh$plot)

results_weekly_sm$plot

(results_weekly_s$plot | results_weekly_p$plot | results_weekly_g$plot) /
  (results_weekly_em$plot | results_weekly_la$plot | results_weekly_sh$plot)
```

RMSE & MAE

```{r}
# observed deaths
observed_deaths_santa_monica_weekly <- santa_monica_weekly$weekly_incidence
observed_deaths_silverlake_weekly <- silverlake_weekly$weekly_incidence
observed_deaths_pomona_weekly <- pomona_weekly$weekly_incidence
observed_deaths_glendale_weekly <- glendale_weekly$weekly_incidence
observed_deaths_el_monte_weekly <- el_monte_weekly$weekly_incidence
observed_deaths_little_armenia_weekly <- little_armenia_weekly$weekly_incidence
observed_deaths_signal_hill_weekly <- signal_hill_weekly$weekly_incidence

# expected/predicted deaths
expected_deaths_santa_monica_weekly <- results_weekly_sm$expected_deaths
expected_deaths_silverlake_weekly <- results_weekly_s$expected_deaths
expected_deaths_pomona_weekly <- results_weekly_p$expected_deaths
expected_deaths_glendale_weekly <- results_weekly_g$expected_deaths
expected_deaths_el_monte_weekly <- results_weekly_em$expected_deaths
expected_deaths_little_armenia_weekly <- results_weekly_la$expected_deaths
expected_deaths_signal_hill_weekly <- results_weekly_sh$expected_deaths


observed_deaths_weekly <- list(
  santa_monica = observed_deaths_santa_monica_weekly,
  silverlake = observed_deaths_silverlake_weekly,
  pomona = observed_deaths_pomona_weekly,
  glendale = observed_deaths_glendale_weekly,
  el_monte = observed_deaths_el_monte_weekly,
  little_armenia = observed_deaths_little_armenia_weekly,
  signal_hill = observed_deaths_signal_hill_weekly
)

expected_deaths_weekly <- list(
  santa_monica = expected_deaths_santa_monica_weekly,
  silverlake = expected_deaths_silverlake_weekly,
  pomona = expected_deaths_pomona_weekly,
  glendale = expected_deaths_glendale_weekly,
  el_monte = expected_deaths_el_monte_weekly,
  little_armenia = expected_deaths_little_armenia_weekly,
  signal_hill = expected_deaths_signal_hill_weekly
)

# calculate RMSE
calculate_rmse_weekly <- function(observed, expected) {
  sqrt(mean((observed - expected)^2, na.rm = TRUE))
}
rmse_values <- sapply(names(observed_deaths_weekly), function(community) {
  calculate_rmse_weekly(observed_deaths_weekly[[community]], 
                        expected_deaths_weekly[[community]])
})
rmse_weekly_results <- data.frame(rmse = rmse_values)

print(rmse_weekly_results)

# calculate MAE
calculate_mae_weekly <- function(observed, expected) {
  mean(abs(observed - expected), na.rm = TRUE)
}
mae_values <- sapply(names(observed_deaths_weekly), function(community) {
  calculate_mae_weekly(observed_deaths_weekly[[community]], 
                       expected_deaths_weekly[[community]])
})
mae_weekly_fit <- data.frame(mae = mae_values)

print(mae_weekly_fit)
```

### Likelihood ratio test: Gamma vs Exponential distribution

- $H_0$: Reject the null hypothesis that the data follows an exponential distribution. The gamma model (with $alpha \neq 1$ fits significantly better.
- $H_a$: Fail to reject the null hypothesis. The simpler exponential model is enough.

1. If LRT rejects null hypothesis, run KS Gamma test.
2. If LRT does not reject null hypothesis, run KS Exponential test.

```{r}
# communities
communities <- list(
  santa_monica = adj_event_times_santa_monica_weekly,
  silverlake = adj_event_times_silverlake_weekly,
  pomona = adj_event_times_pomona_weekly,
  glendale = adj_event_times_glendale_weekly,
  el_monte = adj_event_times_el_monte_weekly,
  little_armenia = adj_event_times_little_armenia_weekly,
  signal_hill = adj_event_times_signal_hill_weekly
)

# Likelihood Ratio Test (Gamma vs. Exponential)
likelihood_ratio_test <- function(event_times, community_name) {
  interarrival_times <- diff(event_times)
  
  gamma_fit <- fitdist(interarrival_times, "gamma")
  exp_fit <- fitdist(interarrival_times, "exp")
  
  loglik_gamma <- sum(dgamma(interarrival_times, shape = gamma_fit$estimate["shape"],
                             rate = gamma_fit$estimate["rate"], log = TRUE))
  
  loglik_exp <- sum(dexp(interarrival_times, rate = exp_fit$estimate["rate"], log = TRUE))
  
  LR_statistic <- -2 * (loglik_exp - loglik_gamma)
  p_value <- pchisq(LR_statistic, df = 1, lower.tail = FALSE)
  
  cat("\n=== Likelihood Ratio Test for Community:", community_name, "===\n")
  cat("Log-likelihood (Gamma):", loglik_gamma, "\n")
  cat("Log-likelihood (Exponential):", loglik_exp, "\n")
  cat("Likelihood Ratio Statistic:", LR_statistic, "\n")
  cat("P-value:", p_value, "\n")
}

# KS Test for Gamma fit
ks_gamma_test <- function(event_times, community_name) {
  inter_event_times <- diff(event_times)
  
  gamma_fit <- fitdist(inter_event_times, "gamma", method = "mle")
  
  shape <- gamma_fit$estimate["shape"]
  rate <- gamma_fit$estimate["rate"]
  
  ks_test <- ks.test(inter_event_times, "pgamma", shape = shape, rate = rate)
  
  cat("\n=== KS Gamma Test for Community:", community_name, "===\n")
  print(ks_test)
}

# KS Test for Exponential fit
ks_exp_test <- function(event_times, community_name) {
  inter_event_times <- diff(event_times)
  
  exp_fit <- fitdist(inter_event_times, "exp", method = "mle")
  
  rate <- exp_fit$estimate["rate"]
  
  ks_test <- ks.test(inter_event_times, "pexp", rate = rate)
  
  cat("\n=== KS Exponential Test for Community:", community_name, "===\n")
  print(ks_test)
}

# loop through communities
for (community in names(communities)) {
  event_times <- communities[[community]]
  
  likelihood_ratio_test(event_times, community)
  ks_gamma_test(event_times, community)
  ks_exp_test(event_times, community)
}
```

