---
title: "README File"
author: "Lauren Huang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Estimating COVID-19 Deaths in LA Communities through a Hawkes Process Model

This project explores spatio-temporal COVID-19 death trends through exploratory data analysis (EDA) and Hawkes process models which use different methods of parameter estimation. All models share a common data import and cleaning procedure.

The 7 communities in the subset that will be modeled:
- Santa Monica
- Silverlake
- Pomona
- Glendale
- El Monte
- Little Armenia
- Signal Hill

## Project Files

- `MASDS_Thesis_EDA.Rmd`: Initial exploratory data analysis (EDA) on the COVID-19 death data.

- `MASDS_Thesis_stelfi_model.Rmd`: Fits a Hawkes process model on both daily and weekly scale death data for all 7 communities using the package stelfi, which uses maximum likelihood estimation (MLE). The stelfi model estimates 1 excitation rate and 1 decay rate parameter. It provides a baseline for comparison with other models.

- `MASDS_Thesis_MLE_model_no_age.Rmd`: Fits a Hawkes process model on daily scale death data for all 7 communities with 6 excitation rate and 3 decay rate parameters, using maximum likelihood estimation (MLE). Fits models using 2 different methods of past data: full history/all past data and 16-day rolling window. Provides a comparison of models using multiple excitation, decay parameters versus 1 excitation, decay parameter as in the stelfi model. Also provides a baseline for comparison of the effects of age in later models.

- `MASDS_Thesis_MLE_model_with_age.Rmd`: Fits a Hawkes process model on daily scale death data for all 7 communities with 6 excitation rate, 3 decay rate, and 8 age covariate parameters, using maximum likelihood estimation (MLE). Fits models using 2 different methods of past data: full history/all past data and 16-day rolling window.

- `MASDS_Thesis_LSE_model.Rmd`: Fits a Hawkes process model on daily scale death data for all 7 communities with 6 excitation rate, 3 decay rate, and 8 age covariate parameters, using least squares estimation (LSE). Fits models using 2 different methods of past data: full history/all past data and 16-day rolling window.

- `MASDS_Thesis_Bayesian_model_allpast_Santa_Monica.Rmd`, 
`MASDS_Thesis_Bayesian_model_allpast_Silverlake.Rmd`
`MASDS_Thesis_Bayesian_model_allpast_Pomona.Rmd`,
`MASDS_Thesis_Bayesian_model_allpast_Glendale.Rmd`,
`MASDS_Thesis_Bayesian_model_allpast_El_Monte.Rmd`,
`MASDS_Thesis_Bayesian_model_allpast_Little_Armenia.Rmd`,
`MASDS_Thesis_Bayesian_model_allpast_Signal_Hill.Rmd`: Fits a Hawkes process model on daily scale death data with 6 excitation rate, 3 decay rate, and 8 age covariate parameters, using Bayesian inference via the inlabru package. Fits models using full history/all past data. Each file has the same code, each of the 7 communities are saved separately in their own file.

`MASDS_Thesis_Bayesian_model_16window_Santa_Monica.Rmd`, 
`MASDS_Thesis_Bayesian_model_16window_Silverlake.Rmd`
`MASDS_Thesis_Bayesian_model_16window_Pomona.Rmd`,
`MASDS_Thesis_Bayesian_model_16window_Glendale.Rmd`,
`MASDS_Thesis_Bayesian_model_16window_El_Monte.Rmd`,
`MASDS_Thesis_Bayesian_model_16window_Little_Armenia.Rmd`,
`MASDS_Thesis_Bayesian_model_16window_Signal_Hill.Rmd`: Fits a Hawkes process model on daily scale death data with 6 excitation rate, 3 decay rate, and 8 age covariate parameters, using Bayesian inference via the inlabru package. Fits models using the 16-day rolling window approach of past data. Each file has the same code, each of the 7 communities are saved separately in their own file.

- `MASDS_Thesis_MLE_model_byyear_Santa_Monica.Rmd`,
`MASDS_Thesis_MLE_model_byyear_Glendale.Rmd`,
`MASDS_Thesis_MLE_model_byyear_Little_Armenia.Rmd`: Fits a Hawkes process model on daily scale death data with 6 excitation rate, 3 decay rate, and 8 age covariate parameters, using maximum likelihood estimation (MLE). Fits models using 2 different methods of past data: full history/all past data and 16-day rolling window. Separates community death data into 3 chunks by year, 2020, 2021 and 2022. 
The communities were chosen based on having different population sizes/death report counts: Santa Monica (moderate), Glendale (large), Little Armenia (small). This model aims to investigate the residuals and see if non-stationarity trends exist.

## How to Run

Each Rmd file can be run independently after installing the required R packages.

Basic Steps:
1. Download all the data directories, all the `.Rmd` files, and all the corresponding `.txt` files which contain the saved parameter estimates. Save them all in the same parent directory.
2. Open a `.Rmd` file in RStudio.
2. Click `Knit` to run the analysis.

See each individual file to load required packages. 
Example required R packages: `utils`, `dplyr`, `purrr`, `fs`, `stringr`, `lubridate`, `stelfi`, `PtProcess`, `stats`, `tidyr`, `ggplot2`, `foreach`, `parallel`, `purrr`, `inlabru`, `INLA`, `sn`, `rstan`, `brms`, `bayesplot`, `patchwork`, `reshape2`, `sf`, `tmap`, `cartogram`, `MASS`, `fitdistrplus`, `geosphere`, `spData`, `geodist`, `vegan`, `ade4`, `scales`, `patchwork`.

## Example

Each R Markdown file can be run independently and includes all necessary steps to load, clean, and model the COVID-19 death data.

To view the full analysis for a given model (ex. the stelfi model), open the corresponding `.Rmd` file in RStudio and run it.

Please note that the `.txt` files contain saved parameter estimates, save them in the same directory as the model `.Rmd` file when you run the code. Using them will save time when running the model code in the R Markdown files. You will notice that in some files, code that was originally used to run the model and save the parameters is commented out for this reason.

The example below shows all the model files, simply run to get the results.

```{r}
# example necessary packages, may differ between model files
# recommended to run files individually
library(utils)
library(dplyr)
library(purrr)
library(fs)
library(stringr)
library(ggplot2)
library(sf)
library(tmap)
library(cartogram)
library(stelfi)
library(lubridate)
library(hawkes)
library(MASS)
library(fitdistrplus)
library(PtProcess)
library(stats)
library(foreach)
library(sn)
library(geosphere)
library(spData)
library(sf)
library(tidyr)
library(geodist)
library(vegan)
library(ade4)
library(scales)
library(patchwork)
library(parallel)
library(reshape2)

# open and knit the files below in RStudio

# Exploratory Data Analysis
MASDS_Thesis_EDA.Rmd
rmarkdown::render("MASDS_Thesis_EDA.Rmd")

# stelfi model
MASDS_Thesis_stelfi_model.Rmd
rmarkdown::render("MASDS_Thesis_stelfi_model.Rmd")

# MLE model (no Age)
MASDS_Thesis_MLE_model_no_age.Rmd
rmarkdown::render("MASDS_Thesis_MLE_model_no_age.Rmd")

# MLE model (with Age)
MASDS_Thesis_MLE_model_with_age.Rmd
rmarkdown::render("MASDS_Thesis_MLE_model_with_age.Rmd")

# LSE model
MASDS_Thesis_LSE_model.Rmd
rmarkdown::render("MASDS_Thesis_LSE_model.Rmd")

# Bayesian models: all past data approach
MASDS_Thesis_Bayesian_model_allpast_Santa_Monica.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_allpast_Santa_Monica.Rmd")

MASDS_Thesis_Bayesian_model_allpast_Silverlake.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_allpast_Silverlake.Rmd")

MASDS_Thesis_Bayesian_model_allpast_Pomona.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_allpast_Pomona.Rmd")

MASDS_Thesis_Bayesian_model_allpast_Glendale.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_allpast_Glendale.Rmd")

MASDS_Thesis_Bayesian_model_allpast_El_Monte.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_allpast_El_Monte.Rmd")

MASDS_Thesis_Bayesian_model_allpast_Little_Armenia.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_allpast_Little_Armenia.Rmd")

MASDS_Thesis_Bayesian_model_allpast_Signal_Hill.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_allpast_Signal_Hill.Rmd")

# Bayesian models: 16-day rolling window approach
MASDS_Thesis_Bayesian_model_allpast_Santa_Monica.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_16window_Santa_Monica.Rmd")

MASDS_Thesis_Bayesian_model_allpast_Silverlake.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_16window_Silverlake.Rmd")

MASDS_Thesis_Bayesian_model_allpast_Pomona.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_16window_Pomona.Rmd")

MASDS_Thesis_Bayesian_model_allpast_Glendale.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_16window_Glendale.Rmd")

MASDS_Thesis_Bayesian_model_allpast_El_Monte.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_16window_El_Monte.Rmd")

MASDS_Thesis_Bayesian_model_allpast_Little_Armenia.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_16window_Little_Armenia.Rmd")

MASDS_Thesis_Bayesian_model_allpast_Signal_Hill.Rmd
rmarkdown::render("MASDS_Thesis_Bayesian_model_16window_Signal_Hill.Rmd")

# MLE by-year models (testing for non-stationarity)
MASDS_Thesis_MLE_model_byyear_Santa_Monica.Rmd
rmarkdown::render("MASDS_Thesis_MLE_model_byyear_Santa_Monica.Rmd")

MASDS_Thesis_MLE_model_byyear_Glendale.Rmd
rmarkdown::render("MASDS_Thesis_MLE_model_byyear_Glendale.Rmd")

MASDS_Thesis_MLE_model_byyear_Little_Armenia.Rmd
rmarkdown::render("MASDS_Thesis_MLE_model_byyear_Little_Armenia.Rmd")
```

