---
title: "MASDS Thesis"
author: "Lauren Huang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(utils)
library(dplyr)
library(purrr)
library(fs)
library(stringr)
library(ggplot2)
library(sf)
library(tmap)
library(cartogram)
library(stelfi)
library(lubridate)
library(hawkes)
library(MASS)
library(fitdistrplus)
library(PtProcess)
library(stats)
library(foreach)
library(sn)
library(geosphere)
library(spData)
library(sf)
library(tidyr)
library(geodist)
library(vegan)
library(ade4)
library(scales)
library(patchwork)
library(parallel)
library(reshape2)
```

# Read in the Data

```{r}
# parent directory containing relevant folders
parent_dir <- "C:/Users/cupca/Desktop/UCLA/Thesis"

# all folders within the parent directory
folder_list <- dir_ls(path = parent_dir, type = "dir")

# folders within the parent directory
folders <- c("2020", "2021", "2022", "2023")

# construct full paths to the specified directories
full_paths <- file.path(parent_dir, folders)

# a function to read and standardize columns of each csv file
read_and_standardize <- function(file) {
  df <- read.csv(file)
  # convert certain columns to the same format
  columns_to_convert <- c("Cases", "Case.Rate")
  # replace occurrences of -- with 0
  value_to_replace <- "--"
  # convert occurrences of the specific character value in the specified columns to integer 0
  df <- df %>%
  mutate(across(all_of(columns_to_convert), 
                 ~ ifelse(. == value_to_replace, 0, .))) %>%
  mutate(across(all_of(columns_to_convert), 
                 ~ as.numeric(as.character(.))))
  
  # standardize columns: add NA/missing columns if they do not exist
  if (!("Deaths" %in% names(df))) {
    df$Deaths <- NA
  }
  if (!("Death.Rate" %in% names(df))) {
    df$Death.Rate <- NA
  }
  return(df)
}

# read, combine all csv files from all folders
all_covid_data <- full_paths %>%
  # list all csv files in the folder
  map(~ dir_ls(path = .x, regexp = "\\.csv$", type = "file")) %>%
  # flatten list of file paths
  unlist() %>%
  # read, standardize each csv file
  map_df(read_and_standardize, .id = "source_id")

# save combined files to a new csv file
write.csv(all_covid_data, "all_covid_data.csv", row.names = FALSE)
```

## Data Cleaning

```{r}
# define a function that relabels source_id as Year
replace_if_contains <- function(df, substring_to_find, new_value) {
  # replace values containing the substring in the specified column (source_id)
  df <- df %>%
    mutate(source_id = ifelse(str_detect(source_id, substring_to_find), new_value, source_id))
  return(df)
}

all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2020", new_value="2020")
all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2021", new_value="2021")
all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2022", new_value="2022")
all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2023", new_value="2023")

# rename column as Year
names(all_covid_data)[names(all_covid_data) == 'source_id'] <- 'year'

# Date Preprocessing

# extract month
all_covid_data$month <- substr(all_covid_data$date, nchar(all_covid_data$date) - 3, nchar(all_covid_data$date) - 2)

# extract day
all_covid_data$day <- substr(all_covid_data$date, nchar(all_covid_data$date) - 1, nchar(all_covid_data$date))

# create a column containing the full date (month, day, year)
all_covid_data$date <- as.Date(paste(all_covid_data$month, all_covid_data$day, all_covid_data$year), "%m%d%Y")

# drop individual month, day, year columns we no longer need
all_covid_data <- subset(all_covid_data, select = -c(year, month, day))

# create new column called incidence, counting new deaths per day in each community
all_covid_data <- all_covid_data %>%
  arrange(name, date) %>%
  group_by(name) %>%
  mutate(incidence = Deaths - lag(Deaths, default = 0)) %>%
  ungroup()

# replace negative or NA values in the incidence column with 0
all_covid_data <- all_covid_data %>%
  mutate(incidence = ifelse(is.na(incidence) | incidence < 0, 0, incidence))

# replace NA values with 0 for incidence
all_covid_data <- all_covid_data %>%
  mutate(Deaths = ifelse(is.na(Deaths), 0, Deaths))

# convert the date column to Date type
community_data <- all_covid_data %>%
  mutate(date = as.Date(date)) %>%
  filter(date > 0) %>%
  filter(('2020-05-16' < date) & (date < '2023-01-02'))

# replace NA, 0 cases with 0.5 or 1
community_data <- community_data %>%
  mutate(Deaths = ifelse(is.na(Deaths), 0.5, Deaths),
         Deaths = ifelse(Deaths == 0, 0.5, Deaths))

# convert dates to numeric time (days since the start of the data)
start_date <- min(community_data$date, na.rm = TRUE)
community_data <- community_data %>%
  mutate(time = as.numeric(difftime(date, start_date, units = "days")))

# remove rows with invalid times
community_data <- community_data %>%
  filter(!is.na(time))
```

## Daily death incidence

prepare daily incidence data

```{r}
# Santa Monica
santa_monica <- community_data %>%
  filter(name == "City of Santa Monica")

# Silverlake
silverlake <- community_data %>%
  filter(name == "Los Angeles - Silverlake")

# Pomona
pomona <- community_data %>%
  filter(name == "City of Pomona")

# Glendale
glendale <- community_data %>%
  filter(name == "City of Glendale")

# El Monte
el_monte <- community_data %>%
  filter(name == "City of El Monte")

# Little Armenia
little_armenia <- community_data %>%
  filter(name == "Los Angeles - Little Armenia")

# Signal Hill
signal_hill <- community_data %>%
  filter(name == "City of Signal Hill")
```

prepare transformed event times for each community

```{r}
# cannot be simultaneous events, need to transform event times
generate_event_times <- function(time, deaths) {
  if (deaths > 0) {
    # distribute event times uniformly within the same day
    return(time + runif(deaths, min = 0, max = 1))  # random times
  } else {
    return(NULL)  # no deaths on this day
  }
}

# generate event times for each community
event_times <- community_data %>%
  group_by(name) %>%
  summarise(event_times = list(unlist(mapply(generate_event_times, time, incidence))))
# using death incidence instead of cumulative deaths

# Santa Monica
event_times_santa_monica <- event_times %>%
  filter(name == "City of Santa Monica") %>%
  pull(event_times)
# sort and get unique times
event_times_santa_monica <- sort(unlist(event_times_santa_monica))
event_times_santa_monica_unique <- unique(event_times_santa_monica)
adj_event_times_santa_monica <- event_times_santa_monica_unique + 
  seq(0, length(event_times_santa_monica_unique) - 1) * 1e-5

# Silverlake
event_times_silverlake <- event_times %>%
  filter(name == "Los Angeles - Silverlake") %>%
  pull(event_times)
# sort and get unique times
event_times_silverlake <- sort(unlist(event_times_silverlake))
event_times_silverlake_unique <- unique(event_times_silverlake)
adj_event_times_silverlake <- event_times_silverlake_unique + 
  seq(0, length(event_times_silverlake_unique) - 1) * 1e-5


# Pomona
event_times_pomona <- event_times %>%
  filter(name == "City of Pomona") %>%
  pull(event_times)
# sort and get unique times
event_times_pomona <- sort(unlist(event_times_pomona))
event_times_pomona_unique <- unique(event_times_pomona)
adj_event_times_pomona <- event_times_pomona_unique + 
  seq(0, length(event_times_pomona_unique) - 1) * 1e-5


# Glendale
event_times_glendale <- event_times %>%
  filter(name == "City of Glendale") %>%
  pull(event_times)
# sort and get unique times
event_times_glendale <- sort(unlist(event_times_glendale))
event_times_glendale_unique <- unique(event_times_glendale)
adj_event_times_glendale <- event_times_glendale_unique + 
  seq(0, length(event_times_glendale_unique) - 1) * 1e-5


# El Monte
event_times_el_monte<- event_times %>%
  filter(name == "City of El Monte") %>%
  pull(event_times)
# sort and get unique times
event_times_el_monte <- sort(unlist(event_times_el_monte))
event_times_el_monte_unique <- unique(event_times_el_monte)
adj_event_times_el_monte <- event_times_el_monte_unique + 
  seq(0, length(event_times_el_monte_unique) - 1) * 1e-5

# Little Armenia
event_times_little_armenia <- event_times %>%
  filter(name == "Los Angeles - Little Armenia") %>%
  pull(event_times)
# sort and get unique times
event_times_little_armenia <- sort(unlist(event_times_little_armenia))
event_times_little_armenia_unique <- unique(event_times_little_armenia)
adj_event_times_little_armenia <- event_times_little_armenia_unique + 
  seq(0, length(event_times_little_armenia_unique) - 1) * 1e-5

# Signal Hill
event_times_signal_hill <- event_times %>%
  filter(name == "City of Signal Hill") %>%
  pull(event_times)
# sort and get unique times
event_times_signal_hill <- sort(unlist(event_times_signal_hill))
event_times_signal_hill_unique <- unique(event_times_signal_hill)
adj_event_times_signal_hill <- event_times_signal_hill_unique + 
  seq(0, length(event_times_signal_hill_unique) - 1) * 1e-5
```


# Hawkes Modeling and Analysis

## Adding age covariate using a custom intensity function

```{r}
# age distribution data: yrs 0-9, 10-19, 20-29,...70-79, 80+
age_labels <- c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80+")

# santa monica
age_data_sm <- c(5544, 6775, 9767, 18200, 11240, 12473, 11598, 8768, 5574) # census reporter

# silverlake
age_data_s <- c(3180, 2920, 5040, 5760, 4950, 3240, 2160, 1440, 900) # city data

# pomona
age_data_p <- c(15552, 17071, 22673, 23058, 20584, 18582, 15188, 8959, 3822) # census reporter

# glendale
age_data_g <- c(16839, 20343, 20629, 28874, 26031, 23322, 24892, 17068, 9034) # census reporter

# el monte
age_data_em <- c(10628, 15791, 14855, 12717, 12851, 13337, 12408, 7863, 3332) # census reporter

# little armenia
age_data_la <- c(2790, 2835, 3510, 3600, 2700, 1980, 1530, 1215, 720) # city data

# signal hill
age_data_sh <- c(981, 1240, 1360, 2127, 1782, 1759, 1400, 558, 356) # census reporter

# function to include age covariate in estimation
get_age_distribution <- function(community_name) {
    if (community_name == "Santa_Monica") {
        return(age_data_sm)
    } else if (community_name == "Silverlake") {
        return(age_data_s)
    } else if (community_name == "Pomona") {
        return(age_data_p)
    } else if (community_name == "Glendale") {
        return(age_data_g)
    } else if (community_name == "El_Monte") {
        return(age_data_em)
    } else if (community_name == "Little_Armenia") {
        return(age_data_la)
    } else if (community_name == "Signal_Hill") {
        return(age_data_sh)
    } else {
        stop("Unknown community")
    }
}
```


## Least Squares Estimation (LSE) Method

Note: Code below adapted from the article "Estimating Covid-19 Transmission Time Using Hawkes Point Processes" by Frederic Schoenberg

https://projecteuclid.org/journals/annals-of-applied-statistics/volume-17/issue-4/Estimating-Covid-19-transmission-time-using-Hawkes-point-processes/10.1214/23-AOAS1765.short

### LSE with AGE, 6 Excitation, 3 Decay parameters DAILY

#### using all past data

```{r}
# calculates RMSE
f_deaths_lse_age_6_3_allpast <- function(p, t3, community_name) {
    k <<- k + 1
    if (min(p) < .0000001) return(9e+20)
    if (sum(p[8:10]) > 1) return(9e+20)
    days <- length(t3)
    
    mu <- p[1]
    K <- p[2:7]
    g <- c(p[8:10], 1 - sum(p[8:10]))
    age <- p[11:18]
    
    age_data <- get_age_distribution(community_name)[1:8]
    total_population <- sum(age_data)
    age_weights <- age_data / total_population
    
    expectedn <- rep(0, days)
    
    for (i in 1:days) {
        influence <- 0
        for (j in 1:(i - 1)) {
            time_diff <- i - j
            decay_bin <- min(ceiling(time_diff / (days / 4)), length(g))
            exc_bin <- min(ceiling(j / (days / 6)), length(K))
            influence <- influence + K[exc_bin] * g[decay_bin] * age_weights %*% age * t3[j]
        }
        expectedn[i] <- max(mu + influence, 1e-10)
    }
    ans <- sqrt(mean((expectedn - t3[1:days])^2))
    cat(k, ans, "\n", file = "rmse_log.txt", append = TRUE)
    return(ans)
}

# fit model
fit_lse_age_6_3_allpast <- function(community_name) {
    df_name <- paste0(tolower(community_name))
    df <- get(df_name, envir = .GlobalEnv)
    t3 <<- df$incidence
    days <- length(t3)
    
    k <<- 1
    initial_params <- c(1, rep(0.5, 6), rep(1/4, 3), rep(0.01, 8))
    
    b1 <- optim(initial_params, f_deaths_lse_age_6_3_allpast, t3 = t3,
                community_name = community_name,
               method = "Nelder-Mead", control = list(maxit = 1000))
    
    b2 <- optim(b1$par, f_deaths_lse_age_6_3_allpast, t3 = t3,
                community_name = community_name,
               method = "Nelder-Mead", control = list(maxit = 10500))
    
    return(b2$par)
}


communities <- c("Santa_Monica", "Silverlake", "Pomona", "Glendale",
                "El_Monte", "Little_Armenia", "Signal_Hill")

result_lse_age_6_3_old <- matrix(0, nrow = 7, ncol = 18)

# clear existing log
file.create("rmse_log.txt")

# run estimation
sink("result_community_lse_6_3_age_old.txt")
sink()
for (m in 1:7) {
    result_lse_age_6_3_old[m,] = fit_lse_age_6_3_allpast(communities[m])
    sink("result_community_lse_6_3_age_old.txt", append = T)
    # append to file
    cat(m, " ", result_lse_age_6_3_old[m,], "\n")
    sink()
}

x1 = scan("C:/Users/cupca/Desktop/UCLA/Thesis/result_community_lse_6_3_age_old.txt") 
x2 <- matrix(x1, ncol = 19, nrow = 7, byrow = TRUE)
result_lse_age_6_3_old <- x2[1:7, 2:19]
lse_6_3_age_coefficients_allpast <- result_lse_age_6_3_old[, 11:18]
```

```{r}
# compares expected vs. observed deaths
compute_expected_deaths_lse_age_6_3_allpast <- function(params, df, community_name) {
    mu <- params[1] 
    K <- params[2:7]  # excitation params
    g <- c(params[8:10], 1 - sum(params[8:10]))  # decay params
    age_params <- params[11:18]  # age coefficients
    
    # age distribution
    age_data <- get_age_distribution(community_name)[1:8]
    total_population <- sum(age_data)
    age_weights <- age_data / total_population 
    age_contrib <- sum(age_weights * age_params)
    
    days <- nrow(df)
    expected_deaths <- numeric(days)
    
    for (i in 1:days) {
        influence <- 0
        for (j in 1:(i - 1)) {
            time_diff <- i - j
            
            # decay bin (based on how far in the past j is from i)
            decay_bin <- min(ceiling(time_diff / (days / 4)), 4)  # 4 bins

            # excitation bin (based on when the past event j occurred)
            exc_bin <- min(ceiling(j / (days / 6)), 6)  # 6 bins
            
            influence <- influence + K[exc_bin] * g[decay_bin] * age_contrib * df$incidence[j]
        }
        expected_deaths[i] <- max(mu + influence, 1e-10)
    }

    return(expected_deaths)
}
```

```{r}
lse_age_6_3_allpast_plot_sm <- ggplot(santa_monica, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3_allpast(result_lse_age_6_3_old[1, ], 
                                            santa_monica, "Santa_Monica"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Santa Monica",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_allpast_plot_s <- ggplot(silverlake, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3_allpast(result_lse_age_6_3_old[1, ], 
                                            silverlake, "Silverlake"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Silverlake",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_allpast_plot_p <- ggplot(pomona, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3_allpast(result_lse_age_6_3_old[1, ], 
                                            pomona, "Pomona"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Pomona",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_allpast_plot_g <- ggplot(glendale, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3_allpast(result_lse_age_6_3_old[1, ], 
                                            glendale, "Glendale"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Glendale",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_allpast_plot_em <- ggplot(el_monte, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3_allpast(result_lse_age_6_3_old[1, ], el_monte, 
                                            "El_Monte"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for El Monte",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_allpast_plot_la <- ggplot(little_armenia, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3_allpast(result_lse_age_6_3_old[1, ], 
                                            little_armenia, 
                                            "Little_Armenia"), color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Little Armenia",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_allpast_plot_sh <- ggplot(signal_hill, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3_allpast(result_lse_age_6_3_old[1, ], signal_hill, 
                                            "Signal_Hill"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Signal Hill",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

# combine plots
(lse_age_6_3_allpast_plot_sm | lse_age_6_3_allpast_plot_s | lse_age_6_3_allpast_plot_p) / 
  (lse_age_6_3_allpast_plot_g | lse_age_6_3_allpast_plot_em | lse_age_6_3_allpast_plot_la) / 
  (lse_age_6_3_allpast_plot_sh)

# Santa Monica
lse_age_6_3_allpast_plot_sm

# other communities
(lse_age_6_3_allpast_plot_s | lse_age_6_3_allpast_plot_p | lse_age_6_3_allpast_plot_g) /
  (lse_age_6_3_allpast_plot_em | lse_age_6_3_allpast_plot_la | lse_age_6_3_allpast_plot_sh)
```

```{r}
# calculate RMSE
compute_rmse <- function(observed, expected) {
    sqrt(mean((observed - expected)^2))
}
rmse_age_lse_values_old <- numeric(7)
for (i in 1:7) {
    expected_deaths <- compute_expected_deaths_lse_age_6_3_allpast(result_lse_age_6_3_old[i, ], 
                                                             get(tolower(communities[i])), 
                                                             communities[i])
    observed_deaths <- get(tolower(communities[i]))$incidence
    rmse_age_lse_values_old[i] <- compute_rmse(observed_deaths, expected_deaths)
}

data.frame(Community = communities, RMSE = rmse_age_lse_values_old)

# calculate MAE
compute_mae <- function(observed, expected) {
  mean(abs(observed - expected), na.rm = TRUE)
}
mae_age_lse_values_old <- numeric(7)
for (i in 1:7) {
    expected_deaths <- compute_expected_deaths_lse_age_6_3_allpast(result_lse_age_6_3_old[i, ], 
                                                             get(tolower(communities[i])), 
                                                             communities[i])
    observed_deaths <- get(tolower(communities[i]))$incidence
    mae_age_lse_values_old[i] <- compute_mae(observed_deaths, expected_deaths)
}

data.frame(Community = communities, MAE = mae_age_lse_values_old)
```


#### 16 day rolling window approach

```{r}
# calculates RMSE
f_deaths_lse_age_6_3 <- function(p, community_name) {
    k <<- k + 1
    cat(k, " ")
    
    if (min(p) < .0000001) return(9e+20)
    if (sum(p[8:10]) > 1) return(9e+20)
    
    mu <- p[1]
    K <- p[2:7] # excitation parameters
    g <- c(p[8:10], 1-sum(p[8:10])) # decay parameters
    age <- p[11:18] # age coefficients
    
    # age distribution
    age_data <- get_age_distribution(community_name)[1:8]
    total_population <- sum(age_data)
    age_weights <- age_data / total_population
    
    # initialize expected deaths
    expectedn <- rep(0, days)
    
    for (i in 1:days) {
        # determine past 16 days of incidence
        if (i > 16) tmp <- t3[i - c(1:16)]
        if (i < 17) tmp <- c(t3[(i-1):1], rep(0, (17-i)))
        if (i < 2) tmp <- rep(0, 16)
        
        # adjust influence of past deaths based on age weights
        ind <- min(ceiling(i / 16), length(K))  # prevent exceeding length of K
        expectedn[i] <- max(mu + sum(K[ind] * g * age_weights * age * tmp), 1e-10)
    }
    
    ans <- sqrt(mean((expectedn - t3[1:days])^2))  # RMSE
    cat(ans, "\n")
    return(ans)
}


# fit model
fit_lse_age_6_3 <- function(community_name) {
    df_name <- paste0(tolower(community_name))
    df <- get(df_name, envir = .GlobalEnv)
    t3 <<- df$incidence
    days <- length(t3)
    
    k <<- 1
    initial_params <- c(1, rep(0.5, 6), rep(1/4, 3), rep(0.01, 8))
    
    b1 <- optim(initial_params, f_deaths_lse_age_6_3, community_name = community_name,
               method = "Nelder-Mead", control = list(maxit = 10000))
    
    b2 <- optim(b1$par, f_deaths_lse_age_6_3, community_name = community_name,
               method = "Nelder-Mead", control = list(maxit = 10000))
    
    return(b2$par)
}

communities <- c("Santa_Monica", "Silverlake", "Pomona", "Glendale",
                "El_Monte", "Little_Armenia", "Signal_Hill")

result_lse_age_6_3 <- matrix(0, nrow = 7, ncol = 18)

# run estimation
sink("result_community_lse_6_3_age.txt")
sink()
for (m in 1:7) {
    result_lse_age_6_3[m,] = fit_lse_age_6_3(communities[m])
    # append to file
    sink("result_community_lse_6_3_age.txt", append=T)
    cat(m, " ", result_lse_age_6_3[m,], "\n")
    sink()
}

x1 = scan("C:/Users/cupca/Desktop/UCLA/Thesis/result_community_lse_6_3_age.txt") 
x2 <- matrix(x1, ncol = 19, nrow = 7, byrow = TRUE)
result_lse_age_6_3 <- x2[1:7, 2:19]
lse_6_3_age_coefficients <- result_lse_age_6_3[, 11:18]
```

```{r}
# compares expected vs. observed deaths
compute_expected_deaths_lse_age_6_3 <- function(params, df, community_name) {
    mu <- params[1]
    K <- params[2:7] # excitation params
    g <- c(params[8:10], 1 - sum(params[8:10])) # decay params
    age_params <- params[11:18]  # age coefficients
    
    # age distribution
    age_data <- get_age_distribution(community_name)[1:8]
    total_population <- sum(age_data)
    age_weights <- age_data / total_population
    
    days <- nrow(df)
    expected_deaths <- numeric(days)
    
    for (i in 1:days) {
        # determine past 16 days of incidence
        if (i > 16) {
            tmp <- df$incidence[(i-16):(i-1)]
        } else {
            tmp <- c(df$incidence[1:(i-1)], rep(0, 16 - (i-1)))  # Pad with 0s if not enough past values
        }
        ind <- min(ceiling(i / 16), length(K))  # Ensure index does not exceed length of K
        
        # adjust influence of past deaths based on age weights
        expected_deaths[i] <- max(mu + sum(K[ind] * g * age_weights * age_params * tmp), 1e-10)
    }
    return(expected_deaths)
}
```

```{r}
lse_age_6_3_plot_sm <- ggplot(santa_monica, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3(result_lse_age_6_3[1, ], 
                                            santa_monica, "Santa_Monica"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Santa Monica",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_plot_s <- ggplot(silverlake, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3(result_lse_age_6_3[1, ], 
                                            silverlake, "Silverlake"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Silverlake",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_plot_p <- ggplot(pomona, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3(result_lse_age_6_3[1, ], 
                                            pomona, "Pomona"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Pomona",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_plot_g <- ggplot(glendale, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3(result_lse_age_6_3[1, ], 
                                            glendale, "Glendale"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Glendale",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_plot_em <- ggplot(el_monte, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3(result_lse_age_6_3[1, ], el_monte, 
                                            "El_Monte"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for El Monte",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_plot_la <- ggplot(little_armenia, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3(result_lse_age_6_3[1, ], 
                                            little_armenia, 
                                            "Little_Armenia"), color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Little Armenia",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

lse_age_6_3_plot_sh <- ggplot(signal_hill, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_lse_age_6_3(result_lse_age_6_3[1, ], signal_hill, 
                                            "Signal_Hill"), 
                color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Signal Hill",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

# combine plots
(lse_age_6_3_plot_sm | lse_age_6_3_plot_s | lse_age_6_3_plot_p) / 
  (lse_age_6_3_plot_g | lse_age_6_3_plot_em | lse_age_6_3_plot_la) / 
  (lse_age_6_3_plot_sh)

# Santa Monica
lse_age_6_3_plot_sm

# other communities
(lse_age_6_3_plot_s | lse_age_6_3_plot_p | lse_age_6_3_plot_g) /
  (lse_age_6_3_plot_em | lse_age_6_3_plot_la | lse_age_6_3_plot_sh)
```

```{r}
# calculate RMSE
compute_rmse <- function(observed, expected) {
    sqrt(mean((observed - expected)^2))
}
rmse_age_lse_6_3_values <- numeric(7)
for (i in 1:7) {
    expected_deaths <- compute_expected_deaths_lse_age_6_3(result_lse_age_6_3[1, ], 
                                                           get(tolower(communities[i])), 
                                                           communities[i])
    observed_deaths <- get(tolower(communities[i]))$incidence
    rmse_age_lse_6_3_values[i] <- compute_rmse(observed_deaths, expected_deaths)
}

data.frame(Community = communities, RMSE = rmse_age_lse_6_3_values)

# calculate MAE
compute_mae <- function(observed, expected) {
  mean(abs(observed - expected), na.rm = TRUE)
}
mae_age_lse_6_3_values <- numeric(7)
for (i in 1:7) {
    expected_deaths <- compute_expected_deaths_lse_age_6_3(result_lse_age_6_3[1, ],
                                                           get(tolower(communities[i])), 
                                                           communities[i])
    observed_deaths <- get(tolower(communities[i]))$incidence
    mae_age_lse_6_3_values[i] <- compute_mae(observed_deaths, expected_deaths)
}

data.frame(Community = communities, MAE = mae_age_lse_6_3_values)
```


