---
title: "MASDS Thesis - MLE model with age"
author: "Lauren Huang"
date: "`r Sys.Date()`"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

```{r}
library(utils)
library(dplyr)
library(purrr)
library(fs)
library(stringr)
library(ggplot2)
library(sf)
library(tmap)
library(cartogram)
library(stelfi)
library(lubridate)
library(hawkes)
library(MASS)
library(fitdistrplus)
library(PtProcess)
library(stats)
library(foreach)
library(sn)
library(geosphere)
library(spData)
library(tidyr)
library(geodist)
library(vegan)
library(ade4)
library(scales)
library(patchwork)
library(parallel)
library(reshape2)
```

# Read in the Data

```{r}
# parent directory containing relevant folders
parent_dir <- "C:/Users/cupca/Desktop/UCLA/Thesis"

# all folders within the parent directory
folder_list <- dir_ls(path = parent_dir, type = "dir")

# folders within the parent directory
folders <- c("2020", "2021", "2022", "2023")

# construct full paths to the specified directories
full_paths <- file.path(parent_dir, folders)

# a function to read and standardize columns of each csv file
read_and_standardize <- function(file) {
  df <- read.csv(file)
  # convert certain columns to the same format
  columns_to_convert <- c("Cases", "Case.Rate")
  # replace occurrences of -- with 0
  value_to_replace <- "--"
  # convert occurrences of the specific character value in the specified columns to integer 0
  df <- df %>%
  mutate(across(all_of(columns_to_convert), 
                 ~ ifelse(. == value_to_replace, 0, .))) %>%
  mutate(across(all_of(columns_to_convert), 
                 ~ as.numeric(as.character(.))))
  
  # standardize columns: add NA/missing columns if they do not exist
  if (!("Deaths" %in% names(df))) {
    df$Deaths <- NA
  }
  if (!("Death.Rate" %in% names(df))) {
    df$Death.Rate <- NA
  }
  return(df)
}

# read, combine all csv files from all folders
all_covid_data <- full_paths %>%
  # list all csv files in the folder
  map(~ dir_ls(path = .x, regexp = "\\.csv$", type = "file")) %>%
  # flatten list of file paths
  unlist() %>%
  # read, standardize each csv file
  map_df(read_and_standardize, .id = "source_id")

# save combined files to a new csv file
write.csv(all_covid_data, "all_covid_data.csv", row.names = FALSE)
```

## Data Cleaning

```{r}
# define a function that relabels source_id as Year
replace_if_contains <- function(df, substring_to_find, new_value) {
  # replace values containing the substring in the specified column (source_id)
  df <- df %>%
    mutate(source_id = ifelse(str_detect(source_id, substring_to_find), new_value, source_id))
  return(df)
}

all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2020", new_value="2020")
all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2021", new_value="2021")
all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2022", new_value="2022")
all_covid_data <- replace_if_contains(df=all_covid_data, substring_to_find="2023", new_value="2023")

# rename column as Year
names(all_covid_data)[names(all_covid_data) == 'source_id'] <- 'year'

# Date Preprocessing

# extract month
all_covid_data$month <- substr(all_covid_data$date, nchar(all_covid_data$date) - 3, nchar(all_covid_data$date) - 2)

# extract day
all_covid_data$day <- substr(all_covid_data$date, nchar(all_covid_data$date) - 1, nchar(all_covid_data$date))

# create a column containing the full date (month, day, year)
all_covid_data$date <- as.Date(paste(all_covid_data$month, all_covid_data$day, all_covid_data$year), "%m%d%Y")

# drop individual month, day, year columns we no longer need
all_covid_data <- subset(all_covid_data, select = -c(year, month, day))

# create new column called incidence, counting new deaths per day in each community
all_covid_data <- all_covid_data %>%
  arrange(name, date) %>%
  group_by(name) %>%
  mutate(incidence = Deaths - lag(Deaths, default = 0)) %>%
  ungroup()

# replace negative or NA values in the incidence column with 0
all_covid_data <- all_covid_data %>%
  mutate(incidence = ifelse(is.na(incidence) | incidence < 0, 0, incidence))

# replace NA values with 0 for incidence
all_covid_data <- all_covid_data %>%
  mutate(Deaths = ifelse(is.na(Deaths), 0, Deaths))

# convert the date column to Date type
community_data <- all_covid_data %>%
  mutate(date = as.Date(date)) %>%
  filter(date > 0) %>%
  filter(('2020-05-16' < date) & (date < '2023-01-02'))
# exclude date gap between 11/21/22-1/2/23 and 1/3/23-11/21/23

# replace NA, 0 cases with 0.5 or 1
community_data <- community_data %>%
  mutate(Deaths = ifelse(is.na(Deaths), 0.5, Deaths),
         Deaths = ifelse(Deaths == 0, 0.5, Deaths))

# convert dates to numeric time (days since the start of the data)
start_date <- min(community_data$date, na.rm = TRUE)
community_data <- community_data %>%
  mutate(time = as.numeric(difftime(date, start_date, units = "days")))

# remove rows with invalid times
community_data <- community_data %>%
  filter(!is.na(time))
```

## Daily death incidence

prepare daily incidence data

```{r}
# Santa Monica
santa_monica <- community_data %>%
  filter(name == "City of Santa Monica")

# Silverlake
silverlake <- community_data %>%
  filter(name == "Los Angeles - Silverlake")

# Pomona
pomona <- community_data %>%
  filter(name == "City of Pomona")

# Glendale
glendale <- community_data %>%
  filter(name == "City of Glendale")

# El Monte
el_monte <- community_data %>%
  filter(name == "City of El Monte")

# Little Armenia
little_armenia <- community_data %>%
  filter(name == "Los Angeles - Little Armenia")

# Signal Hill
signal_hill <- community_data %>%
  filter(name == "City of Signal Hill")
```

prepare transformed event times for each community

```{r}
# cannot be simultaneous events, need to transform event times
generate_event_times <- function(time, deaths) {
  if (deaths > 0) {
    # distribute event times uniformly within the same day
    return(time + runif(deaths, min = 0, max = 1))  # random times
  } else {
    return(NULL)  # no deaths on this day
  }
}

# generate event times for each community
event_times <- community_data %>%
  group_by(name) %>%
  summarise(event_times = list(unlist(mapply(generate_event_times, time, incidence))))
# using death incidence instead of cumulative deaths

# Santa Monica
event_times_santa_monica <- event_times %>%
  filter(name == "City of Santa Monica") %>%
  pull(event_times)
# sort and get unique times
event_times_santa_monica <- sort(unlist(event_times_santa_monica))
event_times_santa_monica_unique <- unique(event_times_santa_monica)
adj_event_times_santa_monica <- event_times_santa_monica_unique + 
  seq(0, length(event_times_santa_monica_unique) - 1) * 1e-5

# Silverlake
event_times_silverlake <- event_times %>%
  filter(name == "Los Angeles - Silverlake") %>%
  pull(event_times)
# sort and get unique times
event_times_silverlake <- sort(unlist(event_times_silverlake))
event_times_silverlake_unique <- unique(event_times_silverlake)
adj_event_times_silverlake <- event_times_silverlake_unique + 
  seq(0, length(event_times_silverlake_unique) - 1) * 1e-5

# Pomona
event_times_pomona <- event_times %>%
  filter(name == "City of Pomona") %>%
  pull(event_times)
# sort and get unique times
event_times_pomona <- sort(unlist(event_times_pomona))
event_times_pomona_unique <- unique(event_times_pomona)
adj_event_times_pomona <- event_times_pomona_unique + 
  seq(0, length(event_times_pomona_unique) - 1) * 1e-5

# Glendale
event_times_glendale <- event_times %>%
  filter(name == "City of Glendale") %>%
  pull(event_times)
# sort and get unique times
event_times_glendale <- sort(unlist(event_times_glendale))
event_times_glendale_unique <- unique(event_times_glendale)
adj_event_times_glendale <- event_times_glendale_unique + 
  seq(0, length(event_times_glendale_unique) - 1) * 1e-5

# El Monte
event_times_el_monte<- event_times %>%
  filter(name == "City of El Monte") %>%
  pull(event_times)
# sort and get unique times
event_times_el_monte <- sort(unlist(event_times_el_monte))
event_times_el_monte_unique <- unique(event_times_el_monte)
adj_event_times_el_monte <- event_times_el_monte_unique + 
  seq(0, length(event_times_el_monte_unique) - 1) * 1e-5

# Little Armenia
event_times_little_armenia <- event_times %>%
  filter(name == "Los Angeles - Little Armenia") %>%
  pull(event_times)
# sort and get unique times
event_times_little_armenia <- sort(unlist(event_times_little_armenia))
event_times_little_armenia_unique <- unique(event_times_little_armenia)
adj_event_times_little_armenia <- event_times_little_armenia_unique + 
  seq(0, length(event_times_little_armenia_unique) - 1) * 1e-5

# Signal Hill
event_times_signal_hill <- event_times %>%
  filter(name == "City of Signal Hill") %>%
  pull(event_times)
# sort and get unique times
event_times_signal_hill <- sort(unlist(event_times_signal_hill))
event_times_signal_hill_unique <- unique(event_times_signal_hill)
adj_event_times_signal_hill <- event_times_signal_hill_unique + 
  seq(0, length(event_times_signal_hill_unique) - 1) * 1e-5
```


# Hawkes Modeling and Analysis

## Adding age covariate using a custom intensity function

```{r}
# age distribution data: yrs 0-9, 10-19, 20-29,...70-79, 80+
age_labels <- c("0-9", "10-19", "20-29", "30-39", "40-49", "50-59", "60-69", "70-79", "80+")

# santa monica
age_data_sm <- c(5544, 6775, 9767, 18200, 11240, 12473, 11598, 8768, 5574) # census reporter

# silverlake
age_data_s <- c(3180, 2920, 5040, 5760, 4950, 3240, 2160, 1440, 900) # city data

# pomona
age_data_p <- c(15552, 17071, 22673, 23058, 20584, 18582, 15188, 8959, 3822) # census reporter

# glendale
age_data_g <- c(16839, 20343, 20629, 28874, 26031, 23322, 24892, 17068, 9034) # census reporter

# el monte
age_data_em <- c(10628, 15791, 14855, 12717, 12851, 13337, 12408, 7863, 3332) # census reporter

# little armenia
age_data_la <- c(2790, 2835, 3510, 3600, 2700, 1980, 1530, 1215, 720) # city data

# signal hill
age_data_sh <- c(981, 1240, 1360, 2127, 1782, 1759, 1400, 558, 356) # census reporter

# function to include age covariate in estimation
get_age_distribution <- function(community_name) {
    if (community_name == "Santa_Monica") {
        return(age_data_sm)
    } else if (community_name == "Silverlake") {
        return(age_data_s)
    } else if (community_name == "Pomona") {
        return(age_data_p)
    } else if (community_name == "Glendale") {
        return(age_data_g)
    } else if (community_name == "El_Monte") {
        return(age_data_em)
    } else if (community_name == "Little_Armenia") {
        return(age_data_la)
    } else if (community_name == "Signal_Hill") {
        return(age_data_sh)
    } else {
        stop("Unknown community")
    }
}
```

### MLE (with Age covariate), 6 Excitation, 3 Decay parameters DAILY

#### uses all past data

```{r}
# list of communities
communities <- list(
  "Santa Monica" = list(event_times = adj_event_times_santa_monica, age_data = age_data_sm),
  "Silver Lake" = list(event_times = adj_event_times_silverlake, age_data = age_data_s),
  "Pomona" = list(event_times = adj_event_times_pomona, age_data = age_data_p),
  "Glendale" = list(event_times = adj_event_times_glendale, age_data = age_data_g),
  "El Monte" = list(event_times = adj_event_times_el_monte, age_data = age_data_em),
  "Little Armenia" = list(event_times = adj_event_times_little_armenia, age_data = age_data_la),
  "Signal Hill" = list(event_times = adj_event_times_signal_hill, age_data = age_data_sh)
)
```

```{r}
# intensity function
custom_intensity_mle_age <- function(params, t, event_times, age_distribution) {
  mu <- max(params[1] + sum(params[2:9] * age_distribution), 1e-5)
  alpha <- params[10:15]
  beta <- params[16:18]

  excitation <- sapply(t, function(time) {
    prev_events <- event_times[event_times < time]
    if (length(prev_events) == 0) return(0)

    t_diff <- time - prev_events
    event_group <- (seq_along(prev_events) %% 6) + 1
    time_group <- findInterval(t_diff, quantile(t_diff, c(0.33, 0.66))) + 1

    sum(alpha[event_group] * exp(-beta[time_group] * t_diff))
  })

  mu + excitation
}

# log-likelihood function
log_likelihood <- function(params, event_times, age_distribution) {
  log_lik <- 0
  for (i in seq_along(event_times)) {
    t <- event_times[i]
    lambda_t <- custom_intensity_mle_age(params, t, event_times[1:(i-1)], age_distribution)

    if (lambda_t <= 0 || !is.finite(lambda_t)) return(1e10)
    log_lik <- log_lik + log(lambda_t)

    integral <- tryCatch({
      integrate(
        custom_intensity_mle_age,
        lower = ifelse(i == 1, 0, event_times[i-1]),
        upper = t,
        params = params,
        event_times = event_times[1:(i-1)],
        age_distribution = age_distribution,
        subdivisions = 100,
        rel.tol = 1e-4
      )$value
    }, error = function(e) {
      warning("Integration failed at event ", i, ": ", e$message)
      return(1e10)
    })

    log_lik <- log_lik - integral
  }
  return(-log_lik)
}
 
# function that saves results to txt file
# save_mle_age_results <- function(results, community, filename = "result_community_mle_6_3_age_daily.txt") {
#   sink(filename, append = file.exists(filename))
# 
#   cat("=== COMMUNITY:", community, "===\n")
#   cat("TIMESTAMP:", format(Sys.time()), "\n")
# 
#   # parameter estimates
#   cat("\n[PARAMETERS]\n")
#   cat("BASELINE:", results$baseline, "\n")
#   cat("AGE_COEFFS:", paste(round(results$age_coefficients, 6), collapse = " "), "\n")
#   cat("EXCITATION_PARAMS:", paste(round(results$excitation_params, 6), collapse = " "), "\n")
#   cat("DECAY_PARAMS:", paste(round(results$decay_params, 6), collapse = " "), "\n")
# 
#   # model diagnostics
#   cat("\n[DIAGNOSTICS]\n")
#   cat("LOGLIK:", results$loglik, "\n")
#   cat("CONVERGENCE:", results$convergence, "\n")
#   cat("N_EVENTS:", results$n_events, "\n")
#   cat("FUNCTION_EVALS:", results$counts[1], "\n")
# 
#   cat("\n---END---\n\n")
#   sink()
# }
# 
# # initial parameters
# initial_params <- c(
#   0.1,              # baseline
#   rep(0.01, 8),     # age coefficients
#   rep(0.1, 6),      # 6 excitation parameters
#   c(0.5, 0.2, 0.1)  # 3 decay parameters
# )
# 
# # create fresh results file
# if (file.exists("result_community_mle_6_3_age_daily.txt")) file.remove("result_community_mle_6_3_age_daily.txt")
# 
# # loops through communities and estimates parameters
# for (community_name in names(communities)) {
#   cat("\n=== Processing", community_name, "===\n")
# 
#   # prepare community data
#   community_data <- communities[[community_name]]
#   age_dist <- community_data$age_data[-9]/sum(community_data$age_data[-9])
# 
#   # run optimization
#   fit <- optim(
#     par = initial_params,
#     fn = log_likelihood,
#     event_times = community_data$event_times,
#     age_distribution = age_dist,
#     method = "L-BFGS-B",
#     lower = c(rep(0, 9), rep(0, 6), rep(0.01, 3)),
#     upper = c(rep(Inf, 9), rep(10, 6), rep(10, 3)),
#     control = list(trace = 1, maxit = 300)
#   )
# 
#   # prepare results
#   mle_age_results_daily <- list(
#     baseline = fit$par[1],
#     age_coefficients = fit$par[2:9],
#     excitation_params = fit$par[10:15],
#     decay_params = fit$par[16:18],
#     loglik = -fit$value,
#     convergence = fit$convergence,
#     n_events = length(community_data$event_times),
#     counts = fit$counts
#   )
# 
#   # save results
#   save_mle_age_results(mle_age_results_daily, community_name)
# 
#   # summary
#   cat("\nCompleted", community_name, "with convergence", fit$convergence, "\n")
#   cat("Log-likelihood:", -fit$value, "\n")
# }
```

```{r}
# function to parse the results file and get estimates
read_mle_age_results <- function(filename = "result_community_mle_6_3_age_daily_old.txt") {

  file_content <- readLines(filename)

  communities <- list()
  current_community <- NULL

  for (line in file_content) {

    # new community
    if (grepl("^=== COMMUNITY:", line)) {
      current_community <- gsub("=== COMMUNITY: (.*?) ===", "\\1", line)
      communities[[current_community]] <- list()
      next
    }

    # skip empty lines and separators
    if (grepl("^---END---|^$", line)) next

    # extract baseline
    if (grepl("^BASELINE:", line)) {
      communities[[current_community]]$baseline <- as.numeric(gsub("BASELINE: (.*)", "\\1", line))
    }

    # extract age coefficients
    if (grepl("^AGE_COEFFS:", line)) {
      communities[[current_community]]$age_coefficients <-
        as.numeric(strsplit(gsub("AGE_COEFFS: (.*)", "\\1", line), " ")[[1]])
    }

    # extract excitation parameters
    if (grepl("^EXCITATION_PARAMS:", line)) {
      communities[[current_community]]$excitation_params <-
        as.numeric(strsplit(gsub("EXCITATION_PARAMS: (.*)", "\\1", line), " ")[[1]])
    }

    # extract decay parameters
    if (grepl("^DECAY_PARAMS:", line)) {
      communities[[current_community]]$decay_params <-
        as.numeric(strsplit(gsub("DECAY_PARAMS: (.*)", "\\1", line), " ")[[1]])
    }

    # extract log-likelihood
    if (grepl("^LOGLIK:", line)) {
      communities[[current_community]]$loglik <- as.numeric(gsub("LOGLIK: (.*)", "\\1", line))
    }
  }
  return(communities)
}

mle_age_results_daily <- read_mle_age_results()

# get all community names
community_names <- names(mle_age_results_daily)

# create a data frame of all baselines
baseline_df <- data.frame(
  Community = community_names,
  Baseline = sapply(mle_age_results_daily, function(x) x$baseline)
)

# create a matrix of all excitation parameters
excitation_matrix <- t(sapply(mle_age_results_daily, function(x) x$excitation_params))
colnames(excitation_matrix) <- paste0("Excitation_", 1:6)

# combine all parameters into one data frame
full_mle_age_results <- do.call(rbind, lapply(names(mle_age_results_daily), function(comm) {
  data.frame(
    Community = comm,
    Baseline = mle_age_results_daily[[comm]]$baseline,
    Age_Coeffs = I(list(mle_age_results_daily[[comm]]$age_coefficients)),
    Excitation = I(list(mle_age_results_daily[[comm]]$excitation_params)),
    Decay = I(list(mle_age_results_daily[[comm]]$decay_params)),
    LogLik = mle_age_results_daily[[comm]]$loglik
  )
}))

print(baseline_df)
print(excitation_matrix)
head(full_mle_age_results)
```

```{r}
# list of all communities
community_data_mle_age <- list(
  "Santa Monica" = list(
    df = santa_monica,
    adj_times = adj_event_times_santa_monica,
    age_data = age_data_sm
  ),
  "Silver Lake" = list(
    df = silverlake,
    adj_times = adj_event_times_silverlake,
    age_data = age_data_s
  ),
  "Pomona" = list(
    df = pomona,
    adj_times = adj_event_times_pomona,
    age_data = age_data_p
  ),
  "Glendale" = list(
    df = glendale,
    adj_times = adj_event_times_glendale,
    age_data = age_data_g
  ),
  "El Monte" = list(
    df = el_monte,
    adj_times = adj_event_times_el_monte,
    age_data = age_data_em
  ),
  "Little Armenia" = list(
    df = little_armenia,
    adj_times = adj_event_times_little_armenia,
    age_data = age_data_la
  ),
  "Signal Hill" = list(
    df = signal_hill,
    adj_times = adj_event_times_signal_hill,
    age_data = age_data_sh
  )
)

process_and_plot_all <- function(community_name, data, adj_event_times, estimates, age_data) {

  # convert event times to dates and numeric days
  event_times <- data$date
  start_date <- min(event_times)
  numeric_event_times <- as.numeric(difftime(event_times, start_date, units = "days"))
  numeric_mapped_times <- as.numeric(difftime(start_date + adj_event_times, start_date, units = "days"))

  # initialize expected deaths
  expected_deaths <- numeric(length(numeric_event_times))

  # calculate expected deaths
  for (i in seq_along(numeric_event_times)) {
    t <- numeric_event_times[i]

    # get previous events
    prev_events <- numeric_mapped_times[numeric_mapped_times < t]

    # calculate intensity
    expected_deaths[i] <- custom_intensity_mle_age(
      params = estimates,
      t = t,
      event_times = prev_events,
      age_distribution = age_data[-9]/sum(age_data[-9]) # exclude total age group
    )
  }

  # results dataframe
  results_df <- data.frame(
    date = event_times,
    observed = data$incidence,
    expected = expected_deaths
  )

  # plot
  p <- ggplot(results_df, aes(x = date)) +
    geom_line(aes(y = observed, color = "Observed"), linewidth = 0.8, alpha = 1) +
    geom_line(aes(y = expected, color = "Expected"), linewidth = 0.8, alpha = 1) +
    scale_color_manual(values = c("Observed" = "black", "Expected" = "red")) +
    labs(title = paste("Expected vs Observed Deaths in ", community_name),
         x = "Date", y = "Daily Deaths",
         color = "Type") +
    theme_minimal() +
    theme(legend.position = "right")

  return(list(
    data = results_df,
    plot = p
  ))
}

generate_all_plots <- function(results_list, community_data) {
  all_plots <- list()

  for (comm in names(community_data_mle_age)) {
    if (!comm %in% names(results_list)) {
      warning("No results found for ", comm, ", skipping...")
      next
    }
    cat("Processing", comm, "...\n")

    all_plots[[comm]] <- process_and_plot_all(
      community_name = comm,
      data = community_data_mle_age[[comm]]$df,
      adj_event_times = community_data_mle_age[[comm]]$adj_times,
      estimates = c(
        results_list[[comm]]$baseline,
        results_list[[comm]]$age_coefficients,
        results_list[[comm]]$excitation_params,
        results_list[[comm]]$decay_params
      ),
      age_data = community_data_mle_age[[comm]]$age_data
    )
  }
  return(all_plots)
}


mle_age_results_daily <- read_mle_age_results("result_community_mle_6_3_age_daily_old.txt")

# generate plots
all_mle_age_results_daily <- generate_all_plots(mle_age_results_daily, community_data_mle_age)

# combine plots
(all_mle_age_results_daily$`Santa Monica`$plot |
    all_mle_age_results_daily$`Silver Lake`$plot |
    all_mle_age_results_daily$Pomona$plot) /
  (all_mle_age_results_daily$Glendale$plot |
     all_mle_age_results_daily$`El Monte`$plot |
     all_mle_age_results_daily$`Little Armenia`$plot) /
  all_mle_age_results_daily$`Signal Hill`$plot

# Santa Monica
all_mle_age_results_daily$`Santa Monica`$plot

# other communities
(all_mle_age_results_daily$`Silver Lake`$plot |
    all_mle_age_results_daily$Pomona$plot |
    all_mle_age_results_daily$Glendale$plot) /
  (all_mle_age_results_daily$`El Monte`$plot |
     all_mle_age_results_daily$`Little Armenia`$plot |
     all_mle_age_results_daily$`Signal Hill`$plot)
```

```{r}
# calculate RMSE, MAE
calculate_metrics <- function(observed, predicted) {
  list(
    RMSE = sqrt(mean((observed - predicted)^2, na.rm = TRUE)),
    MAE = mean(abs(observed - predicted), na.rm = TRUE)
  )
}

# loop through communities
community_metrics <- lapply(names(all_mle_age_results_daily), function(comm) {
  df <- all_mle_age_results_daily[[comm]]$data
  metrics <- calculate_metrics(df$observed, df$expected)

  data.frame(
    Community = comm,
    RMSE = metrics$RMSE,
    MAE = metrics$MAE,
    stringsAsFactors = FALSE
  )
})

metrics_df <- do.call(rbind, community_metrics)

print(metrics_df, row.names = FALSE)
```

```{r}
# spatial data
geodata <- read_sf("C:/Users/cupca/Desktop/UCLA/Thesis/geospatial_data/CHP_2023_Data.shp") %>%
  dplyr::select(-"Geo_ID", -"Label") %>%
  rename("name" = "Geo_Short_")

community_name_map <- c("Santa Monica" = "Santa Monica", 
                        "Silverlake" = "Silverlake",
                        "Pomona" = "Pomona",
                        "Glendale" = "Glendale",
                        "El Monte" = "El Monte",
                        "Little Armenia" = "Little Armenia",
                        "Signal Hill" = "Signal Hill")

# calculate expected deaths
expected_deaths_df <- data.frame(
  name = unname(community_name_map),
  expected_deaths = sapply(names(community_data_mle_age), function(comm) {
    results_df <- all_mle_age_results_daily[[comm]]$data
    sum(results_df$expected, na.rm = TRUE)
  })
)

# merge spatial and expected death data
geodata_subset <- geodata %>%
  filter(name %in% expected_deaths_df$name) %>%
  left_join(expected_deaths_df, by = "name")

# plot
ggplot(geodata_subset) +
  geom_sf(aes(fill = expected_deaths), color = "white") +
  geom_sf_text(aes(label = name), size = 2, color = "black") +
  scale_fill_viridis_c(option = "plasma", na.value = "grey80",
                       name = "Expected Deaths") +
  labs(title = "All past data age MLE model estimated total COVID-19 deaths") +
  theme_minimal()
```

```{r}
geodata <- read_sf("C:/Users/cupca/Desktop/UCLA/Thesis/geospatial_data/CHP_2023_Data.shp")
geodata <- geodata %>% 
  dplyr::select(-"Geo_ID", -"Label") %>% 
  rename("name" = "Geo_Name") # drop unecessary columns

# join expected deaths to geospatial data
geodata_expected <- left_join(geodata, expected_deaths_df, by = "name")

library(ggplot2)

ggplot(geodata_expected) +
  geom_sf(aes(fill = expected_deaths), color = "white") +
  scale_fill_viridis_c(option = "magma", name = "Expected Deaths", trans = "log") +
  labs(title = "Expected COVID-19 Deaths by LA Community based on Age 6 excitation, 3 decay MLE model (all past data)") +
  theme_minimal()
```


#### 16 day rolling window approach

Note: Code below adapted from the article "Estimating Covid-19 Transmission Time Using Hawkes Point Processes" by Frederic Schoenberg

https://projecteuclid.org/journals/annals-of-applied-statistics/volume-17/issue-4/Estimating-Covid-19-transmission-time-using-Hawkes-point-processes/10.1214/23-AOAS1765.short

```{r}
set.seed(85)

# subset community names
communities <- c("Santa_Monica", "Silverlake", "Pomona", "Glendale", 
                 "El_Monte", "Little_Armenia", "Signal_Hill")

# log-likelihood function
log_likelihood <- function(p, community_name, lambda = 0.01) {
    mu <- p[1] # background rate
    K <- p[2:7] # excitation params
    g <- c(p[8:10], 1 - sum(p[8:10]))  # decay params
    age <- p[11:18]  # age coefficients
    
    # age distribution
    age_data <- get_age_distribution(community_name)[1:8]
    total_population <- sum(age_data)
    age_weights <- age_data / total_population
    
    likelihood <- 0
    
    # calculate the conditional intensity of the next event
    for (i in 1:days) {
        # Determine past 16 days of incidence
        if (i > 16) tmp <- t3[i - c(1:16)]
        if (i < 17) tmp <- c(t3[(i-1):1], rep(0, (17-i)))
        if (i < 2) tmp <- rep(0, 16)
        
        tmp <- rev(tmp)
        tmp <- tmp[1:length(age_weights)]
        
        ind <- min(ceiling(i / 16), length(K))

        expectedn <- max(mu + sum(K[ind] * g * age_weights * age * tmp), 1e-10)
        
        if (expectedn <= 0) {
          expectedn <- 1e-10
          }
        # log Poisson likelihood for event counts
        likelihood <- likelihood + dpois(t3[i], lambda = expectedn, log = TRUE)
    }
    age_penalty <- sum(age^2) * lambda  # lambda = tuning parameter (ex 0.01)
    return(-likelihood + age_penalty)
}

# optim to maximize the log-likelihood
fit_daily_mle_age <- function(community_name) {
    df_name <- paste0(tolower(community_name))
    df <- get(df_name, envir = .GlobalEnv)
    t3 <<- df$incidence
    days <- length(t3)
    
    k <<- 1
    initial_params <- c(1, rep(0.5, 6), rep(1/4, 3), rep(1, 8))
    
    b1 <- optim(initial_params, log_likelihood, community_name = community_name,
                    method = "Nelder-Mead", control = list(maxit = 10000))
    
    b2 <- optim(b1$par, log_likelihood, community_name = community_name,
               method = "Nelder-Mead", control = list(maxit = 10000))
    
    return(b2$par)
}

# store results for each community
result_mle_age_daily_6_3 <- matrix(0, ncol = 18, nrow = 7)

# clear file to start fresh before appending new results
# sink("result_community_mle_6_3_age_daily.txt")
# sink()
# for (m in 1:7) {
#     result_mle_age_daily_6_3[m,] = fit_daily_mle_age(communities[m])
#     # append to file
#     sink("result_community_mle_6_3_age_daily.txt", append=T)
#     cat(m, " ", result_mle_age_daily_6_3[m,], "\n")
#     sink()
# }

x1 = scan("C:/Users/cupca/Desktop/UCLA/Thesis/result_community_mle_6_3_age_daily.txt") 
x2 <- matrix(x1, ncol = 19, nrow = 7, byrow = TRUE)
result_mle_age_daily_6_3 <- x2[1:7, 2:19]
mle_6_3_age_coefficients <- result_mle_age_daily_6_3[, 11:18]
```

```{r, warning = F}
# compares expected vs. observed deaths
compute_expected_deaths_age_mle_6_3 <- function(params, df, community_name) {
    mu <- params[1]
    K <- params[2:7] # excitation params
    g <- c(params[8:10], 1 - sum(params[8:10])) # decay params
    age_params <- params[11:18]  # age coefficients
    
    # age distribution
    age_data <- get_age_distribution(community_name)[1:8]
    total_population <- sum(age_data)
    age_weights <- age_data / total_population
    
    days <- nrow(df)
    expected_deaths <- numeric(days)
    
    for (i in 1:days) {
        # Determine past 16 days of incidence
        if (i > 16) {
            tmp <- df$incidence[(i-16):(i-1)]
        } else {
            tmp <- c(df$incidence[1:(i-1)], rep(0, 16 - (i-1)))  # Pad with 0s if not enough past values
        }
        ind <- min(ceiling(i / 16), length(K))  # Ensure index does not exceed length of K
        
        expected_deaths[i] <- max(mu + sum(K[ind] * g * age_weights * age_params * tmp), 1e-10)
    }
    return(expected_deaths)
}
```

```{r}
mle_age_daily_plot_sm <- ggplot(santa_monica, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_age_mle_6_3(result_mle_age_daily_6_3[1, ], santa_monica,
                                            "Santa_Monica"), color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Santa Monica",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()
  
mle_age_daily_plot_s <- ggplot(silverlake, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_age_mle_6_3(result_mle_age_daily_6_3[1, ], silverlake,
                                            "Silverlake"), color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Silverlake",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

mle_age_daily_plot_p <- ggplot(pomona, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_age_mle_6_3(result_mle_age_daily_6_3[1, ], pomona,
                                            "Pomona"), color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Pomona",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

mle_age_daily_plot_g <- ggplot(glendale, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_age_mle_6_3(result_mle_age_daily_6_3[1, ], glendale,
                                            "Glendale"), color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Glendale",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

mle_age_daily_plot_em <- ggplot(el_monte, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_age_mle_6_3(result_mle_age_daily_6_3[1, ], el_monte,
                                            "El_Monte"), color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for El Monte",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

mle_age_daily_plot_la <- ggplot(little_armenia, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_age_mle_6_3(result_mle_age_daily_6_3[1, ], little_armenia,
                                            "Little_Armenia"), color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Little Armenia",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

mle_age_daily_plot_sh <- ggplot(signal_hill, aes(x = date)) +
  geom_line(aes(y = incidence, color = "Observed"), linewidth = 1) +
  geom_line(aes(y = compute_expected_deaths_age_mle_6_3(result_mle_age_daily_6_3[1, ], signal_hill,
                                            "Signal_Hill"), color = "Expected"), linewidth = 1) +
  scale_color_manual(values = c("red", "black")) +
  labs(title = "Expected vs. Observed Deaths for Signal Hill",
       x = "Date", y = "Daily Deaths", color = "Legend") +
  theme_minimal()

# combine plots
(mle_age_daily_plot_sm | mle_age_daily_plot_s | mle_age_daily_plot_p) / 
  (mle_age_daily_plot_g | mle_age_daily_plot_em | mle_age_daily_plot_la) / 
  (mle_age_daily_plot_sh)

# Santa Monica
mle_age_daily_plot_sm

# other communities
(mle_age_daily_plot_s | mle_age_daily_plot_p | mle_age_daily_plot_g) /
  (mle_age_daily_plot_em | mle_age_daily_plot_la | mle_age_daily_plot_sh)
```

```{r}
# calculate RMSE
compute_rmse <- function(observed, expected) {
    sqrt(mean((observed - expected)^2))
}
rmse_mle_age_daily_values <- numeric(7)
for (i in 1:7) {
    expected_deaths <- compute_expected_deaths_age_mle_6_3(result_mle_age_daily_6_3[1, ], 
                                                           get(tolower(communities[i])),
                                                           communities[i])
    observed_deaths <- get(tolower(communities[i]))$incidence
    rmse_mle_age_daily_values[i] <- compute_rmse(observed_deaths, expected_deaths)
}

data.frame(Community = communities, RMSE = rmse_mle_age_daily_values)

# calculate MAE
compute_mae <- function(observed, expected) {
  mean(abs(observed - expected), na.rm = TRUE)
}
mae_mle_age_daily_values <- numeric(7)
for (i in 1:7) {
    expected_deaths <- compute_expected_deaths_age_mle_6_3(result_mle_age_daily_6_3[1, ], 
                                                           get(tolower(communities[i])),
                                                           communities[i])
    observed_deaths <- get(tolower(communities[i]))$incidence
    mae_mle_age_daily_values[i] <- compute_mae(observed_deaths, expected_deaths)
}

data.frame(Community = communities, MAE = mae_mle_age_daily_values)
```

```{r}
# spatial data
geodata <- read_sf("C:/Users/cupca/Desktop/UCLA/Thesis/geospatial_data/CHP_2023_Data.shp") %>%
  dplyr::select(-Geo_ID, -Label) %>%
  rename(name = Geo_Short_)

geodata <- geodata %>%
  mutate(
    clean_name = case_when(
      name == "Santa Monica" ~ "Santa_Monica",
      name == "Silverlake" ~ "Silverlake",
      name == "Pomona" ~ "Pomona",
      name == "Glendale" ~ "Glendale",
      name == "El Monte" ~ "El_Monte",
      name == "Little Armenia" ~ "Little_Armenia",
      name == "Signal Hill" ~ "Signal_Hill",
      TRUE ~ NA_character_
    )
  )

community_data_list <- list(
  Santa_Monica = santa_monica,
  Silverlake = silverlake,
  Pomona = pomona,
  Glendale = glendale,
  El_Monte = el_monte,
  Little_Armenia = little_armenia,
  Signal_Hill = signal_hill
)

# calculate expected total deaths per community
geodata$expected_total <- purrr::map2_dbl(
  geodata$clean_name,
  geodata$name,
  function(clean, full_name) {
    if (is.na(clean) || is.null(community_data_list[[clean]])) return(NA_real_)
    df <- community_data_list[[clean]]
    expected <- compute_expected_deaths_age_mle_6_3(result_mle_age_daily_6_3[1, ], df, clean)
    sum(expected, na.rm = TRUE)
  }
)

# zoom in
geodata_zoom <- geodata %>%
  filter(!is.na(expected_total))

# plot
ggplot(geodata_zoom) +
  geom_sf(aes(fill = expected_total), color = "white") +
  geom_sf_text(aes(label = name), size = 2, color = "black") +
  scale_fill_viridis_c(option = "plasma", na.value = "grey80") +
  labs(title = "16-day rolling window age MLE model estimated COVID-19 deaths",
       fill = "Expected Deaths") +
  theme_minimal()
```

